---
title: "Ecosystem Service Flood Control | Norway"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
---

```{css, echo=FALSE}
table {
  float: left;
}
```

Magnus Merkle (SSB), with Marte Lindseth (SSB), Erik Stange (NINA), Lavinia Svae (SSB) and Trine Randen (SSB) 

Naturregnskapsprosjekt, 23. September 2026

\

# Changelog

Changes since version 7:

-   Changed population input from 2020 to an interpolation between 2020 and 2025
-   Minor fix in LCM category translation

# Abstract and Limitation

In this notebook we quantify the ecosystem service of flood control for the Norwegian mainland. Most operations are executed with functions from R packages dplyr and terra. The notebook is created on the basis of the following sources:

-   Vallecillo et al. (2019) Ecosystem services accounting Part II Pilot accounts for crop and timber provision, global climate regulation and flood control. European Commission: JRC Technical Reports.
-   EC (2023) Guidance note to account for flood control ecosystem service - 1st Draft. Agenda Item 6, Task Force on Ecosystem Accuonting, 28-29 November 2023.
-   Immerzeel & Korkou (2024) Evaluation of the INCA-Tool for the purpose of ecosystem services accounting in Norway. EU Grants Project 2022-NO-EGD- Work Package 2. Norwegian Institute for Nature Research

The quantification process is implemented with reference to the work done by Harry Hodgins, Statistics Ireland. Harry has quantified ES flood control for Ireland using R. We do not copy the Statitics Ireland script in every aspect, but we use it as inspiration for the approach implemented here.

The quantification process follows the EC guidance note. It is important to note that the objective of the process is to identify the contribution of different ecosystems to the provision of flood control (i.e. reduced flood risk) and how this provision changes over time. The process is not designed to provide accurate predictions of the spatial pattern of flood risk and flood risk mitigation at fine spatial resolution. It is not a dynamic hydrological model, but rather an accounting exercise. Therefore, the resulting maps and tables should be interpreted with care and with the limitations of the process in mind. They do not provide accurate spatial accounts of the probability of flood damage.

# 1. Filepaths and Functions

The main functions we use are part of dplyr and terra.
```{r, results = 'hide', message = FALSE, warning=FALSE}
# make space
rm(list = ls())
# packages
library(dplyr)
library(tidyr)
library(terra)
library(sf)
library(knitr)
library(viridisLite)
library(here)
# filepath
dfp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "data","INCA")
# file path to results
rfp<-file.path(here::here(),"Norway_R","output_maps")
# A function to extract levels vectors from factor data stored in rasters 
# rst is raster file, lrn is layer name
extr_lvl_from_raster_lyr<-function(rst,lrn){
  if(is.factor(rst[[lrn]])){
    v1 <- as.vector(values(rst[[lrn]]))
    v2 <- as.vector(levels(rst[[lrn]]))[[1]]
    v3 <- v2$value[v1]
    return(v3)
  } else {
    return(as.vector(values(rst[[lrn]])))
  }
}
gc()
```

# 2. Data Sources

Here is a table of data sources used in this project.
```{r}
dfs<-data.frame(matrix(nrow=10,ncol=8))
colnames(dfs) <- c("Input to","Data","Name","Provider","Resolution","Year","Future Updates","Notes")

dfs[1,]<-c("SPA and SDA",
           "Reference land cover map (LCM) with lookup table",
           "Grunnkart_OT",
           "Kartverket, NIBIO, SSB, Mdir",
           "100m grid",
           "2024",
           "2025",
           "This is probably the most important input because we derive ecosystems, curve numbers, and economic assets from it. Updates should ideally be used to quantify changes in service flows over time. Comparability over time would be relevant for this purpose.")
dfs[2,]<-c("SPA",
           "Hydrological soil type map",
           "LosmasseFlate_20240621 & LosmasseFlate_20240622",
           "Norges Geologiske Undersokelse (NGU)",
           "Vector",
           "2015",
           "Unknown",
           "This map is more precise than the INCA default input.")
dfs[3,]<-c("SPA",
           "Curve Number Lookup Table",
           "curve_number_per_ landcover_soil",
           "Eurostat (INCA default version_2_2)",
           "Not spatial",
           "Probably not relevant",
           "Not required",
           "It would be good with a curve number lookup table for all categories reported in the Norwegian LCM. Then we could avoid the translation into CORINE categories.")
dfs[4,]<-c("SPA",
           "Impervious Density Map",
           "IMD_2018_100m_ eu_03035_V2_0_INCA_ 100m_EPSG3035",
           "Eurostat (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "2018",
           "Unknown",
           "Here as well it might be possible to get hold of better data for the Norwegian context.")
dfs[5,]<-c("SPA and service flow",
           "Slope Map",
           "dt1_slope_INCA_ EPSG3035_ 100m_Byte",
           "European Space Agency (INCA default version_2_2)",
           "100m grid (rescaled)",
           "Probably not relevant",
           "Probably not relevant",
           "We could consider trying a finer resolution slope map in the flow quantification module. To test whether this might make results more accurate. Either using Norwegian elevation data, or by importing the original copernicus data.")
dfs[6,]<-c("SPA",
           "Riparian Zones Map",
           "RiparianZonesEU_ INCA_EPSG3035",
           "EC JRC (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "Unknown",
           "Unknown",
           "If NVE has more accurate data for riparian zones we could try replacing this input. Also: if such zones can change over time, it would be relevant to update this data input over time.")
dfs[7,]<-c("SDA",
           "Flood Plains Map",
           "Flom_AktsomhetOmr",
           "NVE",
           "Vector data (rescaled)",
           "Probably not relevant",
           "Probably not relevant",
           "This is probably the best data we can use for the Norwegian context. Alternatively we could use areas where actual floods have happened.")
dfs[8,]<-c("SDA",
           "Population Map",
           "GHS_POP_EYYYY_ GLOBE_R2023A_ 54009_100_V1_0",
           "GHSL",
           "100m grid (rescaled)",
           "2020 & 2025",
           "Yes (every 5 years)",
           "This is another key input which should be varied over time to quantify changes in service flow.")
dfs[9,]<-c("Service Flow",
           "Elevation Map",
           "dt1_dem_INCA_ EPSG3035_100m",
           "European Space Agency (INCA default version_2_2)",
           "100m grid (rescaled)",
           "Probably not relevant",
           "Probably notrelevant",
           "We could try a finer resolution map to see whether that changes results. We could also include areas outside Norway if cross-border flows are flood-relevant for Norway.")
dfs[10,]<-c("Service Flow",
           "Various shapefiles (municipalities, NUTS 2 regions, catchment areas",
           "Various",
           "Various",
           "Vectors",
           "Not relevant",
           "Not relevant",
           "Only to retrieve results for specific subnational regions.")
#print
kable(dfs)
# clean up
rm(dfs)
```

# 3. Service Provision (SPA)

## 3.1 Overview

The first step in the quantification of flood control is to create a raster map of potential runoff retention. This map is a key input to the ensuing steps. We start from the Norwegian land cover map and a European Hydrological Soil Charateristics map. Each combination of land cover and soil characteristic is associated to a Curve Number (CN) value - an index of infiltration/runoff. The resulting curve number raster map is then further refined using various inputs. Once the runoff retention map is obtained, we derive Service Providing Areas (SPA) from it using specific thresholds.

## 3.2 Reference CN Map

### 3.2.1 LCM

Read in and plot our land cover categories.
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# read Norway Grunnkart (this is our reference LCM) ----------------------------
lcm<-terra::rast(file.path(dfp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))
terra::plot(lcm, main="Norwegian Land Cover Map (Grunnkart)")
```

Now the fact of the matter is that the categories used in the Norwegian LCM do not perfectly correspond to the categories used in the curve number table. Therefore we first need to translate the Norwegian categories into the system used in the CORINE data. Note we treat the sea as NA because  it is not relevant for flood control. 
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# make a numeric version of the lcm --------------------------------------------
lcm2<-lcm
activeCat(lcm2) <- "Value"
lcm2<-as.numeric(lcm2)
# read in grunnkart lookup -----------------------------------------------------
lcm_idf<-read.csv(file.path(dfp,"Norway_files","Grunnkart",
                            "Lookup_grunnkart.csv"))
# read in curve number table ---------------------------------------------------
cn_idf<-read.csv(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                           "Lookup_tables",
                           "curve_number_per_landcover_soil.csv"))
# adjust lcm to feature official CORINE categories -----------------------------
# (otherwise our CN lookup table does not fit)
# I create this concordance using the following sources
# https://land.copernicus.eu/content/corine-land-cover-nomenclature-guidelines/html/
# https://nibio.brage.unit.no/nibio-xmlui/bitstream/handle/11250/3120510/NIBIO_RAPPORT_2024_10_28.pdf?sequence=1&isAllowed=y
lcm_idf['CN_cover_id']<-NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id %in% 
                      c(100,111,112,121,130,131,132,133,134,135,142,152)] <- 1
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 141] <- 141
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 200] <- 241
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 310] <- 231
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 320] <- 321
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 400] <- 312
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 410] <- 311
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 420] <- 312
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 440] <- 313
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 520] <- 322
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 600] <- 333
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 611] <- 332
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 623] <- 333
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 631] <- 335 # NA in CN lookup glaciers and perpetual snow
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 720] <- 322
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 810] <- 511 # NA in cn lookup rivers
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 910] <- 512 # NA in cn lookup inland water bodies
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 1100] <- 331 # NA in cn lookup beaches dunes sands
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 1200] <- NA # NA 523 in cn lookup (sea and ocean) - not needed
# and then we add the last category that is not available (uklassifisert)
lcm_idf <- rbind(lcm_idf,c(NA,9900,"unclassified","uklassifisert",NA))
lcm_idf$landcover_id <- as.numeric(lcm_idf$landcover_id)
lcm_idf$CN_cover_id  <- as.numeric(lcm_idf$CN_cover_id)
# make a new lcm that features the CN_cover_id
# numeric version
lcm4 <- terra::subst(lcm2,lcm_idf$landcover_id,lcm_idf$CN_cover_id)
# categorical version
lcm3<-as.factor(lcm4)
terra::plot(lcm3, main="Grunnkartet translated to CORINE LCM categories")
gc()
```

So this is our reference land cover map. Moving on.

### 3.2.2 Hydrological Soil Type

The INCA default hydrological soil type map is coarse and has missing areas in the north. Here you see it:
```{r,message = FALSE, results = 'hide'}
# read hydrological soil type map ----------------------------------------------
soilsd<-terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "Hydrological_soil_type_map",
                             "hydro_soilgroup_INCA_EPSG3035.tif"))
# mm comment: only four categories seems a bit coarse...
# mm comment: the soils data is masked in a weird way in the north of Norway. 
# I think we should really use something better here.
# first crop and mask soils ----------------------------------------------------
soilsd<-crop(soilsd,lcm)
soilsd<-mask(soilsd,lcm)
terra::plot(soilsd, main="INCA default hydrological soil type masked to Norway")
# mm comment: so here you see again how the soil map is weirdly masked in the 
# north. comes from the input data. To be sorted out later
```

We choose to use a Norway-specific dataset instead, which is more detailed. As this data also has gaps, we make adjustments to it in a later step.
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# read in hydrol soil type data ------------------------------------------------
los1<-terra::vect(file.path(dfp,"Norway_files","Losmasse","Losmasse","LosmasseFlate_20240621.shp"))
los2<-terra::vect(file.path(dfp,"Norway_files","Losmasse","Losmasse","LosmasseFlate_20240622.shp"))
gc()
# reproject
los1p <- terra::project(los1,lcm3)
los2p <- terra::project(los2,lcm3)
los1r<-terra::rasterize(los1p,lcm3,field = "infilt",fun="mean")
los2r<-terra::rasterize(los2p,lcm3,field = "infilt",fun="mean")
# The two maps look like they are complements to each other, so we add them together.
soilsu <- sum(los1r,los2r,na.rm=T)
soilsu <- terra::mask(soilsu,lcm3)
rm(los1,los2,los1p,los2p,los1r,los2r)
# And then the plot
# Generate default viridis colors
mod_colors <- viridis(4)
# Replace category 5 with red
mod_colors[5] <- "red"
terra::plot(soilsu, col=mod_colors,main="Hydrological soil type for Norway",sub="Note category 5 means no data available")
gc()
```

So in the above map everything that is red is missing data. The way we fix is is to first compute the correlation between soil type and land cover, and then fill the gaps (areas with soil type == 5) by the "most likely" soil type, given land cover.
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# first we convert our maps into vectors and bind them in a table
corrdf<-as.data.frame(cbind(values(lcm),values(soilsu)))
# then compute the "most likely" soilype for each land cover
most_likely <- corrdf %>%
  group_by(Class, infilt) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(infilt %in% c(1,2,3,4)) %>%
  group_by(Class) %>%
  mutate(total = sum(count),
         percentage = 100 * count / total) %>%
  slice_max(order_by = count, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(most_likely_soil = infilt)
# so then we make a map with the most likely soil type
mlsoil<-terra::classify(lcm,most_likely[,1:2])
mlsoil[mlsoil %in% c(1200,9900)]<-NA
# and then we replace category 5 in our soils map with the most likely soil type.
soils<-soilsu
soils[soils==5]<-mlsoil[soils==5]
terra::plot(soils, main="Adjusted hydrological soil type for Norway",sub="missing areas filled with estimated values")
rm(soilsd,soilsu,corrdf,mlsoil,most_likely)
gc()
```

Ok this is better.

### 2.2.3 Curve Number

So this is now basically a lookup exercise. My approach is to vectorise the rasters, then run the merge process, then rasterise again. There are probably more efficient ways to do this, but this one is simple and it seems to work.
```{r,message = FALSE, results = 'hide', fig.height = 9}
gc()
# extract data vectors
#lcmv <-extr_lvl_from_raster_lyr(lcm3,"value")
lcmv<-values(lcm3)[,1]
soilsv <-extr_lvl_from_raster_lyr(soils,"infilt")
# put into a dataframe
dfc<-as.data.frame(matrix(NA,nrow=length(lcmv),ncol=2))
colnames(dfc)<-c("lcmd","soilsd")
dfc$lcmd <- as.numeric(lcmv)
dfc$soilsd <- soilsv
# join with CN lookup table
dfc <- dfc %>% left_join(cn_idf,by = join_by(lcmd == landcover_id,soilsd == Soil))
# then make a raster from the vector
cnm <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
values(cnm)<-dfc$CN
# plot result
terra::plot(cnm, range=c(0,100), main="Norway Curve Number Map", sub="High number implies high runoff, low number implies low runoff")
# clean up ---------------------------------------------------------------------
rm(dfc,soils,lcmv,soilsv)
gc()
```

Note the map has a few holes, and this comes from cells that have no CN number in the lookup table. The missing categories are: 9900 (uklassifisert), 1200 (marine økosystemer (uspesifisert)), 1100 (Kyststrender dyner og våtmark (uspesifisert)), 910 (Innsjøer og dammer), 810 (Elver og bekker), and 631 (Isbreer og varige snøfonner). I have checked an alternative data source (https://www.nature.com/articles/s41597-019-0155-x) and found that those categories are also missing there. So it is probably just common to assume no curve number for them. Assuming no CN implies that these cells cannot provide flood control.

## 3.3 CN Map Modification

In the following subsections we make adjustments to the curve number map so it becomes more nuanced. We do this by following the process described in the guidance note. In some cases the equations from the guidance note are faulty, for example, plus and minus signs are mixed up. These faults have been identified and correctd by Statistics Ireland and we follow the corrected version. 

### 3.3.1 Imperviousness Correction

We use the INCA default impervious density map, masked to our reference raster.
```{r,message = FALSE, results = 'hide'}
imp <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "Impervious_density_map",
                             "IMD_2018_100m_eu_03035_V2_0_INCA_100m_EPSG3035.tif"))
# first crop and mask to Norway
imp<-crop(imp,lcm,mask=T)
terra::plot(imp, main="Impervious Density Norway")
```

The modification process for imperviousness follows an equation that comes from the guidance note. Impervious density is given as an index between 0 and 100. The process basically increases the CN number in areas where the ground is impervious (where it does not easily absorb water). We implement this straight with the rasters, and then plot the result. Differences are small and mostly visible in cities. Here zooming in on the Oslo area.
```{r, message = FALSE, results = 'hide', fig.width = 12}
cnmi <- 98*(imp/100) + (1-(imp/100))*cnm
# make a stack
cnm1 <- c(cnm,cnmi)
names(cnm1)<-c("CN","CN Adjusted 1")
# plot for Oslo area
terra::plot(cnm1,range=c(0,100),ext = c(4200000,4400000,4000000,4200000),maxcell=ncell(lcm3))
# clean up
rm(imp,cnmi,cnm)
gc()
```

### 3.3.2 Slope Correction

We use the INCA default DEM and Slope maps, cropped and masked to Norway. The modification process again follows an equation provided in the guidance note. Here first a plot of slopes. Note: 100% slope means 45 degrees incline, 200% slope means 63.4 degrees incline, Inf % slope means 90 degrees incline.
```{r,message = FALSE, results = 'hide', fig.height = 9}
slop <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "DEM_and_Slope_maps",
                             "dt1_slope_INCA_EPSG3035_100m_Byte.tif"))
slop<-crop(slop,lcm)
slop<-mask(slop,lcm)
plot(slop, main="Slope Norway [expressed in %")
```

Our data provides slopes in percent. The equation stated in the guidance note based on Huang et al 2016 requires slopes as ratio (ratio = percent/100). Zero slope is NA in the data, so I first turn all NA to zero before running the process. Then apply the formula as stated in the guidance note.  Note: This adjustment process increases the CN number for cells that have a steep slope. Finally, we make a plot to see what difference the operation makes. All in all the differences are very small and we need to zoom in a lot to see anything. The example area provided below is close to Åndalsnes.
```{r, message = FALSE, results = 'hide', fig.width = 12}
slop[is.na(slop)]<-0
cnma <- cnm1[['CN Adjusted 1']] * ((322.79 + (15.63*slop/100))/((slop/100)+323.52))
cnm2<-c(cnm1,cnma)
names(cnm2)<-c("CN","CN Adjusted 1","CN Adjusted 1_2")
plot(cnm2[[c(2,3)]],range=c(0,100),ext = c(4200000,4210000,4370000,4380000),maxcell=ncell(lcm3))
rm(slop,cnma,cnm1)
gc()
```

### 3.3.3 Riparian Zone Correction

The last adjustment is made for riparian zones. Again we use INCA default data for this adjustment and we follow the process decribed in the guidance note. Here a plot of riparian zones in Norway.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# import data
rip <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                              "Riparian_zones_map",
                              "RiparianZonesEU_INCA_EPSG3035.tif"))


# first crop and mask to Norway
rip<-crop(rip,lcm)
rip<-mask(rip,lcm)
# to make it better visible
rip[is.na(rip) & !is.na(lcm3)]<-0
# plot
plot(rip,main="Riparian Zones Norway")
```

And then the process. It is quite straightforward: If a cell is part of a riparian zone, it receives the lowest possible CN value that exists for the land cover category. So it basically saying: All cells that are riparian zones have hydrological soil type 1 (see section 3.2.2). The process therefore reduces the CN numbers in riparian zones, but it does not change the CN map by much. Below a plot zooming in on just west of Hønefoss.
```{r, message = FALSE, results = 'hide', fig.width = 12}
# define relevant landcover categories as per guidance note
lcm_idf['rip_relevant']<-0
relevant_codes <- c(244,311,312,313,321,322,323,324,411,412,421,422,423)
# mm note: most of those do not exist in our Grunnkart in Norway...
# the only ones in Norway are category 3 (forest)
lcm_idf$rip_relevant[lcm_idf$CN_cover_id %in% relevant_codes]<-1
# now for those cells where we are in a relevant riparian zone we will attribute
# the lowest possible cn number for the respective land cover. 
# get vector of rip
ripv <- extr_lvl_from_raster_lyr(rip,"RiparianZonesEU_INCA_EPSG3035")
# get vector of lcm
lcmv <- values(lcm3)[,1]
lcmv <-as.numeric(lcmv)
# get vector of CN
cnv <- extr_lvl_from_raster_lyr(cnm2,"CN Adjusted 1_2")
# relevant rip lcmv
dfc<-as.data.frame(matrix(NA,ncol=3,nrow=length(lcmv)))
colnames(dfc)<-c("lcm","rip","relevant")
#first the land cover data
dfc$lcm <- lcmv
#then the riparian zone data
dfc$rip <- ripv
# then the relevant cells
dfc$relevant[dfc$rip==1 & dfc$lcm %in% relevant_codes] <- 1
# then find the lowest possible CN value for each cell
dfc <- dfc %>% 
  left_join(cn_idf %>% filter(Soil==1) %>% select(-Soil),by = join_by(lcm == landcover_id)) %>%
  rename(CN1 = CN)
#clean up
rm(rip,ripv,lcmv)
dfc <- dfc %>% select(-c(lcm,rip))
gc()
#then get the cn number for each cell
dfc['CNIS'] <- cnv
rm(cnv)
#adjusted CN is the lowest CN
dfc['CNISR'] <- dfc$CN1
gc()
# apart from in rip-irrelevant cells, where it is the old CN value
dfc$CNISR[is.na(dfc$relevant)] <- dfc$CNIS[is.na(dfc$relevant)]
gc()
# then make a new raster
cnma <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
values(cnma)<-dfc$CNISR
names(cnma)<-c("CN Adjusted 1_2_3")
# then stack with our other 
cnm2 <- c(cnm2,cnma)
# and plot where we can see something
plot(cnm2[[c(3,4)]],range=c(0,100),ext = c(4325000,4335000,4115000,4125000),maxcell=ncell(lcm3))
# clean up
rm(dfc,relevant_codes,cnma)
gc()
```

That was it for adjustments. The exercise might be more precise if we could get hold of more detailed and specific data for Norway.

## 2.4 Runoff Retention Map

We just need to reverse the scale of the CN map for this. So that a high number implies that water is absorbed, and a low number implies that water is not absorbed. This is now the main map from which we derive flood control service provision.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# identify max value
max_value <- as.numeric(global(cnm2[[4]],fun = "max", na.rm=T))
# create map
potm <- max_value - cnm2[[4]]
names(potm)<-"Pot_Runuff_Retention"
# save
writeRaster(potm,file.path(rfp,"runoff_retention.tif"),overwrite=T)
# plot
plot(potm, range=c(0,100),main = "Potential Runoff Retention Map")
# clean up
rm(max_value)
gc()
```

## 3.5 SPA Thresholds

So now based on our potential runoff retention map we define our service providing areas (SPA). There are three different ones, depending on whether the areas are located on urban/artificial land, agricultural land, or natural land. SP areas are binary, so 1/0. In order to define them, we use runoff retention thresholds, one for each parent land category. If a cell has a runoff retention value that is at least as large as the threshold, it is defined as service providing cell. This means that it provides the service of flood control, because it retains water.

Statistics Ireland came up with their own thresholds, calculated the same way as the EU guidance note suggests. For our version I have first tried to use the thresholds that the guidance note suggests for Europe (27 for artificial land, 52 for agricultural land, 61 for natural land), but then found that these may not be so applicable for Norway. One problem was, for example, that none of the Norwegian agricultural land could be classified as flood control service providing area, simply because the runoff retention threshold of 52 is too high (there is no agricultural patch in Norway that exceeds this threshold). It might be that Norwegian agricultural land is more impervious than EU average agricultural land. If we went with EU thresholds, then no Norwegian agricultural land could provide flood control.

I therefore calculate Norway specific thresholds using the approach that the guidance note suggests. However, I think this should be revisited to ensure that the threshold definition process has some more justification than only that it is based on a transparent process. With the current process to define thresholds, the following applies:

-   If variation of runoff retention values increases all else equal, the threshold becomes lower. This increases the likelihood that cells get defined as service providing areas.
-   If the mean of runoff retention values increases all else equal, the threshold becomes higher. This reduces the likelihood that cells get defined as service providing areas.

MM: I see the point of incorporating standard variation in the formula to calculate the thresholds, because this avoids inflated threholds when outliers inflate the mean, and it accounts for heterogeneity. Other than that, however, I am not sure whether this threshold-based definition of service provision comes with desirable properties. I suggest to discuss this further and check other sources than the EU guidance note.
```{r, message = FALSE, results = 'hide'}
## first on artificial
artif_id<-c(1,141)
spam_artif_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_artif_am[[1]]<-0
spam_artif_am[is.na(lcm3)]<-NA
for(i in 1:length(artif_id)){
  spam_artif_am[lcm4 == artif_id[i]]<-mean(potm[lcm4 == artif_id[i]],na.rm=T) + sd(potm[lcm4 == artif_id[i]],na.rm=T)
}
spa_tld_artif <- mean(spam_artif_am[lcm4 %in% artif_id],na.rm=T)
## then on agricultural
agric_id<-c(241,231)
spam_agric_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_agric_am[[1]]<-0
spam_agric_am[is.na(lcm3)]<-NA
for(i in 1:length(agric_id)){
  spam_agric_am[lcm4 == agric_id[i]]<-mean(potm[lcm4 == agric_id[i]],na.rm=T) + sd(potm[lcm4 == agric_id[i]],na.rm=T)
}
spa_tld_agric <- mean(spam_agric_am[lcm4 %in% agric_id],na.rm=T)
## then on natural
natur_id<-c(321,312,311,313,322,333,332)
spam_natur_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_natur_am[[1]]<-0
spam_natur_am[is.na(lcm3)]<-NA
for(i in 1:length(natur_id)){
  spam_natur_am[lcm4 == natur_id[i]]<-mean(potm[lcm4 == natur_id[i]],na.rm=T) - sd(potm[lcm4 == natur_id[i]],na.rm=T)
}
spa_tld_natur <- mean(spam_natur_am[lcm4 %in% natur_id],na.rm=T)
## clean up
rm(spam_artif_am,spam_agric_am,spam_natur_am)
gc()
```
Here are the thresholds.
```{r}
# print thresholds
print(paste("Artificial tld:",spa_tld_artif))
print(paste("Agricultural tld:",spa_tld_agric))
print(paste("Natural tld:",spa_tld_natur))
```


## 3.6 SPA Maps

Now given our thresholds we make a stack of maps for the SPAs
```{r, message = FALSE, results = 'hide', fig.height = 9}
# Artificial spa
spam_artif<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_artif[[1]]<-0
spam_artif[is.na(lcm3)]<-NA
spam_artif[lcm4 %in% artif_id & potm >= spa_tld_artif] <-1
# Agricultural spa
spam_agric<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_agric[[1]]<-0
spam_agric[is.na(lcm3)]<-NA
spam_agric[lcm4 %in% agric_id & potm >= spa_tld_agric] <-1
# Natural spa
spam_natur<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_natur[[1]]<-0
spam_natur[is.na(lcm3)]<-NA
spam_natur[lcm4 %in% natur_id & potm >= spa_tld_natur] <-1
# write raster
writeRaster(spam_artif,file.path(rfp,"spam_artif.tif"),overwrite=T)
writeRaster(spam_agric,file.path(rfp,"spam_agric.tif"),overwrite=T)
writeRaster(spam_natur,file.path(rfp,"spam_natur.tif"),overwrite=T)
# stack together and plot
spam <-c(spam_artif,spam_agric,spam_natur)
names(spam)<-c("SPA Artificial","SPA Agricultural", "SPA Natural")
ccspam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(spam[[1]],col=ccspam,colNA="grey40",main = "SPA Artificial")
plot(spam[[2]],col=ccspam,colNA="grey40",main = "SPA Agricultural")
plot(spam[[3]],col=ccspam,colNA="grey40",main = "SPA Natural")
# clean up
rm(ccspam,spam_artif,spam_agric,spam_natur)
gc()
```

# 4. Service Demanding Areas (SDA)

## 4.1 Calculation and Results

SDAs are service demanding areas. These are defined as economic assets located in flood risk areas. Formally: total spatial extent of economic assets that are located in an area which NVE defines as "aktsomhetsområde". Our definition differs from the EU definition, in that we use flood risk areas, whereas the EU uses areas that get flooded at least once within a specific horizon of time. The second kind of service demanding areas are total amount of population inhabiting flood risk areas. 

So first, we start reading in the data and making it fit to our reference map. Note that the original NVE flood risk map is a vector file. We first rasterise this file to our default grid, and we do this outside of this markdown script because it is a time-intensive operation. The rasterisation algorithm defines cells as flood risk areas if their center is within a flood risk vector shape.

Here are our flood risk areas.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# population
# first 2020
popr20<-rast(file.path(dfp,"population","Mollweide","GHS_POP_E2020_GLOBE_R2023A_54009_100_V1_0.tif"))
popr20<-crop(popr20,ext(250000,
                    2000000,
                    6600000,
                    8000000) # 8700000 if svalbard included
                    )
popr20 <- terra::project(popr20,lcm,mask=T)
# then 2025
popr25<-rast(file.path(dfp,"population","Mollweide","GHS_POP_E2025_GLOBE_R2023A_54009_100_V1_0.tif"))
popr25<-crop(popr25,ext(250000,
                    2000000,
                    6600000,
                    8000000) # 8700000 if svalbard included
                    )
popr25 <- terra::project(popr25,lcm,mask=T)
# then the interpolated 2024
popr <- popr20 + ((4/5)*(popr25 - popr20))
rm(popr20,popr25)
# flood
flood <- terra::rast(file.path(dfp,"Norway_files","NVE_flom","NVEKartdata",
                                     "NVEData","Flom_Aktsomhet",
                                     "Flom_AktsomhetOmr_form.tif"))
plot(flood, main = "Flood Risk Areas Norway")
gc()
```

Now it is literally just a question of identifying cells in the flood areas. And then plot results.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# service demand by population
sdam_pop <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_pop[[1]] <- 0
sdam_pop[is.na(lcm3)]<-NA
sdam_pop[flood == 1 & popr>0] <- popr[flood == 1 & popr>0]
# service demand by artificial land
sdam_artif <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_artif[[1]] <- 0
sdam_artif[is.na(lcm3)]<-NA
sdam_artif[flood == 1 & lcm4 %in% artif_id] <- 1
# service demand by agricultural land
sdam_agric <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_agric[[1]] <- 0
sdam_agric[is.na(lcm3)]<-NA
sdam_agric[flood == 1 & lcm4 %in% agric_id] <- 1
# aggregate to economic assets
sdam_econ <- sdam_artif + sdam_agric
# save
writeRaster(sdam_pop,file.path(rfp,"sdam_pop.tif"),overwrite=T)
writeRaster(sdam_econ,file.path(rfp,"sdam_econ.tif"),overwrite=T)
# plot
plot(sdam_econ,col=c("grey90","blue"),main = "Economic assets in flood risk")
ccsdam<-data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey90","blue"))
plot(sdam_pop,col=ccsdam, main = "Population in flood risk")
#clean up
rm(lcm3,lcm4,popr,flood,ccsdam)
gc()
```

Important to note for the next sections: "Economic assets" are defined as cells that are either agricultural land or artificial land. So we add two service demand categories together for this.

## 4.2 Plots for KS Project

First we need to import all KS shapefiles and street data.
```{r, message = FALSE, results = 'hide'}
ahs<-terra::vect(file.path(dfp,"Norway_files","Aurskog","aurskog.shp"))
ahr<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","ahr.shp"))
ehs<-terra::vect(file.path(dfp,"Norway_files","Evje","evje_formatted.shp"))
ehr<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","ehr.shp"))
hos<-terra::vect(file.path(dfp,"Norway_files","Holtalen","holtalen.shp"))
hor<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","hor.shp"))
mos<-terra::vect(file.path(dfp,"Norway_files","Moss","moss.shp"))
mor<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","mor.shp"))
nrs<-terra::vect(file.path(dfp,"Norway_files","Nordreisa","nordreisa.shp"))
nrr<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","nrr.shp"))
sds<-terra::vect(file.path(dfp,"Norway_files","Stad","stad.shp"))
sdr<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","sdr.shp"))
vrs<-terra::vect(file.path(dfp,"Norway_files","Vardo","vardo.shp"))
vrr<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","vrr.shp"))
vas<-terra::vect(file.path(dfp,"Norway_files","Vaagan","vaagan.shp"))
var<-terra::vect(file.path(dfp,"Norway_files","ks_formatted_shapes_wosm","var.shp"))
gc()
```
And then let's plot service demand for our case studies.
```{r, message = FALSE, results = 'hide', fig.height = 7}
# colour code
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
# Aurskog Høland
plot(sdam_econ %>% crop(ahs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Aurskog Holand") 
plot(ahr,add=T)
plot(sdam_pop %>% crop(ahs,mask=T),colNA="grey40",
     main="Population in flood risk, Aurskog Holand") 
plot(ahr,add=T)
# Evje and Hornnes
plot(sdam_econ %>% crop(ehs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Evje and Hornnes") 
plot(ehr,add=T)
plot(sdam_pop %>% crop(ehs,mask=T),colNA="grey40",
     main="Population in flood risk, Evje and Hornnes") 
plot(ehr,add=T)
# Holtålen
plot(sdam_econ %>% crop(hos,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Holtaalen") 
plot(hor,add=T)
plot(sdam_pop %>% crop(hos,mask=T),colNA="grey40",
     main="Population in flood risk, Holtaalen") 
plot(hor,add=T)
#Moss
plot(sdam_econ %>% crop(mos,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Moss") 
plot(mor,add=T)
plot(sdam_pop %>% crop(mos,mask=T),colNA="grey40",
     main="Population in flood risk, Moss") 
plot(mor,add=T)
# Nordreisa
plot(sdam_econ %>% crop(nrs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Nordreisa") 
plot(nrr,add=T)
plot(sdam_pop %>% crop(nrs,mask=T),colNA="grey40",
     main="Population in flood risk, Nordreisa") 
plot(nrr,add=T)
# Stad
plot(sdam_econ %>% crop(sds,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Stad") 
plot(sdr,add=T)
plot(sdam_pop %>% crop(sds,mask=T),colNA="grey40",
     main="Population in flood risk, Stad") 
plot(sdr,add=T)
# Vardø
plot(sdam_econ %>% crop(vrs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Vardo") 
plot(vrr,add=T)
plot(sdam_pop %>% crop(vrs,mask=T),colNA="grey40",
     main="Population in flood risk, Vardo") 
plot(vrr,add=T)
# Vågan
plot(sdam_econ %>% crop(vas,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in flood risk, Vaagan") 
plot(var,add=T)
plot(sdam_pop %>% crop(vas,mask=T),colNA="grey40",
     main="Population in flood risk, Vaagan") 
plot(var,add=T)
# clean up
rm(ccsdam)
gc()
```

# 5. Service Flow SPA --> SDA

## 5.1 Introduction

Now the service flow is quantified at the level of the SDAs. We calculate it for each SDA gridcell as the ratio of upstream surface area that is covered by SPA. Upstream surface area is the area from which we expect water to flow - so this requires slope information.

NB: the guidance suggests to use all upstream area for this, irrespective of political boundaries. For the time being we limit ourselves to Norway as service providing area, but could include neighbouring countries as well if we want.

The calculation of service flow is straightforward using the built-in functions of R's terra package. First, we compute the number of upstream cells for each SDA cell. Second, we compute the proportion of these upstream cells that are covered by SPA (of various kind). The service flow is then equal to the ratio of upstream cells covered by SPA over all upstream cells.

Example: If the the ratio is 1, then all upstream cells of a service demanding cell are covered by SPA, hence "flood controlled by the ecosystem". If the ratio is 0, then none of the upstream cells of a service demanding cell are covered b y SPA, hence "not flood controlled by the ecosystem".

Important to note here is that when there is no unique flow direction to identify from the slope data, then terra will allocate the flow direction randomly. This is not ideal, but at least we can make terra return unique results by setting seeds for the random sequence generator.

## 5.2 Flow Accumulation

First we import elevation data and compute flow direction and flow accumulation for each grid cell.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# import data and plot
elev <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                              "DEM_and_Slope_maps",
                              "dt1_dem_INCA_EPSG3035_100m.tif"))
elev<-crop(elev,lcm)
elev<-mask(elev,lcm)
plot(elev,main="Norway Elevation")
# then the computation
set.seed(2025) # I do this for replicability. Flowdir allocates direction randomly when it is impossible to find a unique direction from the elevation data
flowdir_total <- terra::terrain(elev,v="flowdir")
# plot accumulation
flow_acc_total <- terra::flowAccumulation((flowdir_total))
plot(flow_acc_total, main = "Flow accumulation (number of cells upstream flowing into each cell, example Honefoss",ext = c(4325000,4345000,4105000,4125000),maxcell=ncell(lcm2))
# clean up
rm(elev)
gc()
```

## 5.3 Ratio of Protection Upstream

### 5.3.1 Calculation and Results

Ok so now we tell terra to count upstream cells, but only those that are not SPAs. This gives us a hypothetical flow without water from SPA cells. So we basically simulate that SPA cells retain water. The way I implement this is to use the weight argument in the flowAccummulation function (I have tested the behaviour first on a hypothetical 9x9 grid).

As an alternative version, we also compute a ratio of upstream protection based on a continuous indicator of service provision, which is simply our runoff retention map from section 2.4. The difference this makes is that now all upstream cells can retain some water, and they do so to varying degrees as defined by our runoff retention map. The ratio of upstream protection for a cell is equal to one, if all upstream cells have a runoff retention value of 100 (i.e. a curve number of zero). The ratio is zero if all upstream cells have a runoff retention value of 0 (i.e. a curve number of 100). In most cases, of course, the ratio is somewhere between zero and one.
```{r, message = FALSE, results = 'hide', fig.height = 9}
## ratio artificial spa
tmp<-spam[[1]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_artif<- 1 - flow_acc_tmp/flow_acc_total 
# clean up
rm(tmp,flow_acc_tmp)
## ratio agricultural spa
tmp<-spam[[2]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_agric<- 1 - flow_acc_tmp/flow_acc_total 
# clean up
rm(tmp,flow_acc_tmp)
## ratio nature spa
tmp<-spam[[3]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_natur<- 1 - flow_acc_tmp/flow_acc_total
# clean up
rm(tmp,flow_acc_tmp)
## ratio continuous SPA
tmp<- 1-(potm/100)
# the weight map cannot have NAs. So I turn all NA cells into 1, implying that water
# just flows across them (zero water retention)
tmp[is.na(tmp)]<-1
# then restricted flow accumulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_contn <- 1 - flow_acc_tmp/flow_acc_total 
# need to mask it so the sea becomes NA again
ratio_contn <- mask(ratio_contn,sdam_econ)
# clean up
rm(tmp,flow_acc_tmp)
## plot results
plot(ratio_artif,colNA="grey40", main = "Ratio of upstream cells covered by Artificial SPA")
plot(ratio_agric,colNA="grey40",main = "Ratio of upstream cells covered by Agricultural SPA")
plot(ratio_natur,colNA="grey40",main = "Ratio of upstream cells covered by Natural SPA")
plot(ratio_contn,colNA="grey40",main = "Ratio of upstream runoff retained (continuous service provision)")
## save results as tif
writeRaster(ratio_agric,file.path(rfp,"ratio_agric.tif"),overwrite=T)
writeRaster(ratio_artif,file.path(rfp,"ratio_artif.tif"),overwrite=T)
writeRaster(ratio_natur,file.path(rfp,"ratio_natur.tif"),overwrite=T)
# for ration contn we make a lighter version
ratio_contn_3d <- round(ratio_contn,digits=3)
writeRaster(ratio_contn_3d,file.path(rfp,"ratio_contn_3d.tif"),overwrite=T)
## clean up
#rm(flow_acc_total,flowdir_total)
rm(ratio_contn_1d)
gc()
```

### 5.3.2 Plots for KS Project - natural binary SPA

Here the ratio plots for our KS cases. Only ratio of upstream cells covered by natural SPA.
```{r, message = FALSE, results = 'hide', fig.height = 7}
# Aurskog Høland
plot(ratio_natur %>% crop(ahs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Aurskog Holand") 
plot(ahr,add=T)
# Evje and Hornnes
plot(ratio_natur %>% crop(ehs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Evje and Hornnes") 
plot(ehr,add=T)
# Holtålen
plot(ratio_natur %>% crop(hos,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Holtaalen") 
plot(hor,add=T)
#Moss
plot(ratio_natur %>% crop(mos,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Moss") 
plot(mor,add=T)
# Nordreisa
plot(ratio_natur %>% crop(nrs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Nordreisa") 
plot(nrr,add=T)
# Stad
plot(ratio_natur %>% crop(sds,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Stad") 
plot(sdr,add=T)
# Vardø
plot(ratio_natur %>% crop(vrs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Vardo") 
plot(vrr,add=T)
# Vågan
plot(ratio_natur %>% crop(vas,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Vaagan") 
plot(vas,add=T)
# clean up
gc()
```

### 5.3.3 Plots for KS Project - continuous SPA

Here the ratio plots for our KS cases. Continuous runoff retention across all land covers.

```{r, message = FALSE, results = 'hide', fig.height = 7}
# Aurskog Høland
plot(ratio_contn %>% crop(ahs,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Aurskog Holand") 
plot(ahr,add=T)
# Evje and Hornnes
plot(ratio_contn %>% crop(ehs,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Evje and Hornnes") 
plot(ehr,add=T)
# Holtålen
plot(ratio_contn %>% crop(hos,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Holtaalen") 
plot(hor,add=T)
#Moss
plot(ratio_contn %>% crop(mos,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Moss") 
plot(mor,add=T)
# Nordreisa
plot(ratio_contn %>% crop(nrs,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Nordreisa") 
plot(nrr,add=T)
# Stad
plot(ratio_contn %>% crop(sds,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Stad") 
plot(sdr,add=T)
# Vardø
plot(ratio_contn %>% crop(vrs,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Vardo") 
plot(vrr,add=T)
# Vågan
plot(ratio_contn %>% crop(vas,mask=T),colNA="grey40",
     main="Ratio of upstream runoff retained (continuous service provision), Vaagan") 
plot(vas,add=T)
# clean up
gc()
```

## 5.4 Service Flow

### 5.4.1 Calculation and Results

Now we have the ratios, so next we can compute for each service demanding area what the respective service flow is. We can make maps of it and also tables.
```{r, message = FALSE,results='hide'}
## serice flow to meet demand by artificial area
flow_d_artif_s_artif<-sdam_artif*ratio_artif
flow_d_artif_s_agric<-sdam_artif*ratio_agric
flow_d_artif_s_natur<-sdam_artif*ratio_natur
flow_d_artif_c<-sdam_artif*ratio_contn
## serice flow to meet demand by agricultural area
flow_d_agric_s_artif<-sdam_agric*ratio_artif
flow_d_agric_s_agric<-sdam_agric*ratio_agric
flow_d_agric_s_natur<-sdam_agric*ratio_natur
flow_d_agric_c<-sdam_agric*ratio_contn
## serice flow to meet demand by population
flow_d_pop_s_artif<-sdam_pop*ratio_artif
flow_d_pop_s_agric<-sdam_pop*ratio_agric
flow_d_pop_s_natur<-sdam_pop*ratio_natur
## here add the aggregate ones econ and pop, and save those as rasters.
flow_d_econ <- flow_d_artif_s_artif + flow_d_artif_s_agric + flow_d_artif_s_natur +
  flow_d_agric_s_artif + flow_d_agric_s_agric + flow_d_agric_s_natur
flow_d_pop <- flow_d_pop_s_artif + flow_d_pop_s_agric + flow_d_pop_s_natur
## then the continuous version
flow_d_econ_c <- sdam_econ*ratio_contn
flow_d_pop_c <- sdam_pop*ratio_contn
## total flows
flow_d_all_s_all<-flow_d_artif_s_artif+flow_d_artif_s_agric+flow_d_artif_s_natur+
  flow_d_agric_s_artif+flow_d_agric_s_agric+flow_d_agric_s_natur
## save results as tif
# rfp<-file.path("C:","Users","muk",
#                paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
#                "Projects","Naturregnskap","Flomkontroll","Norway_R","output_maps")
writeRaster(flow_d_econ,file.path(rfp,"flow_d_econ.tif"),overwrite=T)
writeRaster(flow_d_pop,file.path(rfp,"flow_d_pop.tif"),overwrite=T)
writeRaster(flow_d_econ_c,file.path(rfp,"flow_d_econ_c.tif"),overwrite=T)
writeRaster(flow_d_pop_c,file.path(rfp,"flow_d_pop_c.tif"),overwrite=T)
# Make a table of values
flow_df<-as.data.frame(matrix(ncol=4,nrow=5))
colnames(flow_df)<-c("provision_cat","demand_artif","demand_agric","demand_pop")
flow_df$provision_cat<-c("artif","agric","natur","binary_all","contn_all")
# for demand artif
flow_df$demand_artif[flow_df$provision_cat=="artif"]<-sum(values(flow_d_artif_s_artif),na.rm=T)
flow_df$demand_artif[flow_df$provision_cat=="agric"]<-sum(values(flow_d_artif_s_agric),na.rm=T)
flow_df$demand_artif[flow_df$provision_cat=="natur"]<-sum(values(flow_d_artif_s_natur),na.rm=T)
flow_df$demand_artif[flow_df$provision_cat=="binary_all"]<-sum(flow_df[1:3,2])
flow_df$demand_artif[flow_df$provision_cat=="contn_all"]<-sum(values(flow_d_artif_c),na.rm=T)
# for demand agric
flow_df$demand_agric[flow_df$provision_cat=="artif"]<-sum(values(flow_d_agric_s_artif),na.rm=T)
flow_df$demand_agric[flow_df$provision_cat=="agric"]<-sum(values(flow_d_agric_s_agric),na.rm=T)
flow_df$demand_agric[flow_df$provision_cat=="natur"]<-sum(values(flow_d_agric_s_natur),na.rm=T)
flow_df$demand_agric[flow_df$provision_cat=="binary_all"]<-sum(flow_df[1:3,3])
flow_df$demand_agric[flow_df$provision_cat=="contn_all"]<-sum(values(flow_d_agric_c),na.rm=T)
# for demand pop
flow_df$demand_pop[flow_df$provision_cat=="artif"]<-sum(values(flow_d_pop_s_artif),na.rm=T)
flow_df$demand_pop[flow_df$provision_cat=="agric"]<-sum(values(flow_d_pop_s_agric),na.rm=T)
flow_df$demand_pop[flow_df$provision_cat=="natur"]<-sum(values(flow_d_pop_s_natur),na.rm=T)
flow_df$demand_pop[flow_df$provision_cat=="binary_all"]<-sum(flow_df[1:3,4])
flow_df$demand_pop[flow_df$provision_cat=="contn_all"]<-sum(values(flow_d_pop_c),na.rm=T)
#clean up
rm(flow_d_artif_s_artif,flow_d_artif_s_agric,flow_d_artif_s_natur,flow_d_agric_s_artif,flow_d_agric_s_agric,flow_d_agric_s_natur,flow_d_pop_s_artif,flow_d_pop_s_agric,flow_d_pop_s_natur,ratio_agric,ratio_artif,ratio_natur)
gc()
```
Here is the table
```{r}
# print table
kable(flow_df)
```

### 5.4.2 Plots for KS Project - binary SPA

Let's take a look at the flows for our municipalities of interest in the KS project. Only total flow (from all SPAs) to economic assets and to population. Service demanding areas are shaded lighter, so we can compare service flows to demanded service.

```{r, message = FALSE, results = 'hide', fig.height = 6}
# Aurskog Høland
#econ
plot(sdam_econ %>% crop(ahs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Aurskog Holand")
plot(flow_d_econ %>% crop(ahs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(ahr,add=T)
#pop
plot(sdam_pop %>% crop(ahs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Aurskog Holand")
plot(flow_d_pop %>% crop(ahs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(ahr,add=T)
# Evje and Hornnes
#econ
plot(sdam_econ %>% crop(ehs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Evje and Hornnes")
plot(flow_d_econ %>% crop(ehs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(ehr,add=T)
#pop
plot(sdam_pop %>% crop(ehs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to populations, Evje and Hornnes")
plot(flow_d_pop %>% crop(ehs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(ehr,add=T)
# Holtålen
#econ
plot(sdam_econ %>% crop(hos,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Holtaalen")
plot(flow_d_econ %>% crop(hos,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(hor,add=T)
#pop
plot(sdam_pop %>% crop(hos,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Holtaalen")
plot(flow_d_pop %>% crop(hos,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(hor,add=T)
#Moss
#econ
plot(sdam_econ %>% crop(mos,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Moss")
plot(flow_d_econ %>% crop(mos,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(mor,add=T)
#pop
plot(sdam_pop %>% crop(mos,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Moss")
plot(flow_d_pop %>% crop(mos,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(mor,add=T)
# Nordreisa
#econ
plot(sdam_econ %>% crop(nrs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assetss, Nordreisa")
plot(flow_d_econ %>% crop(nrs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(nrr,add=T)
#pop
plot(sdam_pop %>% crop(nrs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Nordreisa")
plot(flow_d_pop %>% crop(nrs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(nrr,add=T)
# Stad
#econ
plot(sdam_econ %>% crop(sds,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assetss, Stad")
plot(flow_d_econ %>% crop(sds,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(sdr,add=T)
#pop
plot(sdam_pop %>% crop(sds,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to populations, Stad")
plot(flow_d_pop %>% crop(sds,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(sdr,add=T)
# Vardø
#econ
plot(sdam_econ %>% crop(vrs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Vardo")
plot(flow_d_econ %>% crop(vrs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(vrr,add=T)
#pop
plot(sdam_pop %>% crop(vrs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Vardo")
plot(flow_d_pop %>% crop(vrs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(vrr,add=T)
# Vågan
#econ
plot(sdam_econ %>% crop(vas,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Vaagan")
plot(flow_d_econ %>% crop(vas,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(var,add=T)
#pop
plot(sdam_pop %>% crop(vas,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Vaagan")
plot(flow_d_pop %>% crop(vas,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(var,add=T)
# clean up
gc()
```

### 5.4.3 Plots for KS Project - continuous SPA

Let's take a look at the flows for our municipalities of interest in the KS project. Only total flow (from all SPAs) to economic assets and to population. Here now based on continuous SPA. Service demanding areas are shaded lighter, so we can compare service flows to demanded service.

```{r, message = FALSE, results = 'hide', fig.height = 6}
# Aurskog Høland
#econ
plot(sdam_econ %>% crop(ahs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Aurskog Holand")
plot(flow_d_econ_c %>% crop(ahs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(ahr,add=T)
#pop
plot(sdam_pop %>% crop(ahs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Aurskog Holand")
plot(flow_d_pop_c %>% crop(ahs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(ahr,add=T)
# Evje and Hornnes
#econ
plot(sdam_econ %>% crop(ehs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Evje and Hornnes")
plot(flow_d_econ_c %>% crop(ehs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(ehr,add=T)
#pop
plot(sdam_pop %>% crop(ehs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to populations, Evje and Hornnes")
plot(flow_d_pop_c %>% crop(ehs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(ehr,add=T)
# Holtålen
#econ
plot(sdam_econ %>% crop(hos,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Holtaalen")
plot(flow_d_econ_c %>% crop(hos,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(hor,add=T)
#pop
plot(sdam_pop %>% crop(hos,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Holtaalen")
plot(flow_d_pop_c %>% crop(hos,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(hor,add=T)
#Moss
#econ
plot(sdam_econ %>% crop(mos,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Moss")
plot(flow_d_econ_c %>% crop(mos,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(mor,add=T)
#pop
plot(sdam_pop %>% crop(mos,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Moss")
plot(flow_d_pop_c %>% crop(mos,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(mor,add=T)
# Nordreisa
#econ
plot(sdam_econ %>% crop(nrs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assetss, Nordreisa")
plot(flow_d_econ_c %>% crop(nrs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(nrr,add=T)
#pop
plot(sdam_pop %>% crop(nrs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Nordreisa")
plot(flow_d_pop_c %>% crop(nrs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(nrr,add=T)
# Stad
#econ
plot(sdam_econ %>% crop(sds,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assetss, Stad")
plot(flow_d_econ_c %>% crop(sds,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(sdr,add=T)
#pop
plot(sdam_pop %>% crop(sds,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to populations, Stad")
plot(flow_d_pop_c %>% crop(sds,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(sdr,add=T)
# Vardø
#econ
plot(sdam_econ %>% crop(vrs,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Vardo")
plot(flow_d_econ_c %>% crop(vrs,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(vrr,add=T)
#pop
plot(sdam_pop %>% crop(vrs,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Vardo")
plot(flow_d_pop_c %>% crop(vrs,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(vrr,add=T)
# Vågan
#econ
plot(sdam_econ %>% crop(vas,mask=T),colNA="white",
     col=data.frame(value=c(1,0),col=c("white","grey10")),legend=F,
     main="Flow to economic assets, Vaagan")
plot(flow_d_econ_c %>% crop(vas,mask=T),colNA="grey40",add=T,alpha=0.8) 
plot(var,add=T)
#pop
plot(sdam_pop %>% crop(vas,mask=T),colNA="white",
     col=data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey10","white")),legend=F,
     main="Flow to population, Vaagan")
plot(flow_d_pop_c %>% crop(vas,mask=T),colNA="grey40",add=T,alpha=0.9) 
plot(var,add=T)
# clean up
gc()
```

# 6. Met and Unmet Demand

## 6.1 Table - binary SPA

Unmet demand is equal to total demand minus provided service. We can express both as absolute and relative numbers. I first make a table for this.
```{r, message = FALSE,results='hide'}
# I make a table of all
dem_df<-as.data.frame(matrix(ncol = 4,nrow=7))
colnames(dem_df)<-c("indicator","demand_artif","demand_agric","demand_pop")
dem_df$indicator<-c("total_demand","total_provided","unmet_demand","rel_prov_artif","rel_prov_agric","rel_prov_natur","rel_unmet")
# total demand
dem_df$demand_artif[dem_df$indicator=="total_demand"]<-sum(values(sdam_artif),na.rm=T)
dem_df$demand_agric[dem_df$indicator=="total_demand"]<-sum(values(sdam_agric),na.rm=T)
dem_df$demand_pop[dem_df$indicator=="total_demand"]<-sum(values(sdam_pop),na.rm=T)
# total provided
dem_df$demand_artif[dem_df$indicator=="total_provided"]<-flow_df$demand_artif[flow_df$provision_cat=="binary_all"]
dem_df$demand_agric[dem_df$indicator=="total_provided"]<-flow_df$demand_agric[flow_df$provision_cat=="binary_all"]
dem_df$demand_pop[dem_df$indicator=="total_provided"]<-flow_df$demand_pop[flow_df$provision_cat=="binary_all"]
# unmet
dem_df$demand_artif[dem_df$indicator=="unmet_demand"]<-dem_df$demand_artif[dem_df$indicator=="total_demand"] - dem_df$demand_artif[dem_df$indicator=="total_provided"]
dem_df$demand_agric[dem_df$indicator=="unmet_demand"]<-dem_df$demand_agric[dem_df$indicator=="total_demand"] - dem_df$demand_agric[dem_df$indicator=="total_provided"]
dem_df$demand_pop[dem_df$indicator=="unmet_demand"]<-dem_df$demand_pop[dem_df$indicator=="total_demand"] - dem_df$demand_pop[dem_df$indicator=="total_provided"]
# relative provision by artificial cells
dem_df$demand_artif[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_artif[flow_df$provision_cat=="artif"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_agric[flow_df$provision_cat=="artif"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_pop[flow_df$provision_cat=="artif"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)
# relative provision by agricultural cells
dem_df$demand_artif[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_artif[flow_df$provision_cat=="agric"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_agric[flow_df$provision_cat=="agric"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_pop[flow_df$provision_cat=="agric"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)
# relative provision by natural cells
dem_df$demand_artif[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_artif[flow_df$provision_cat=="natur"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_agric[flow_df$provision_cat=="natur"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_pop[flow_df$provision_cat=="natur"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)
# relative unmet demand
dem_df$demand_artif[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_artif[dem_df$indicator=="unmet_demand"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_agric[dem_df$indicator=="unmet_demand"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_pop[dem_df$indicator=="unmet_demand"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)
# clean up
gc()
```
Here is the table
```{r}
# print table
kable(dem_df)
```

Relevant to note here is that we use population number in case of demand category "pop" and we use cells in case of demand categories "artif" and "agric". We may want to switch all to the count of cells. Then the respective result just has to be multiplied by the spatial size of each cell (each cell one hectare).

## 6.2 Table - continuous SPA

Unmet demand is equal to total demand minus provided service. We can express both as absolute and relative numbers. I first make a table for this.
```{r, message = FALSE,results='hide'}
# I make a table of all
dem_dfc<-as.data.frame(matrix(ncol = 4,nrow=4))
colnames(dem_dfc)<-c("indicator","demand_artif","demand_agric","demand_pop")
dem_dfc$indicator<-c("total_demand","total_provided","unmet_demand","rel_unmet")
# total demand
dem_dfc$demand_artif[dem_dfc$indicator=="total_demand"]<-sum(values(sdam_artif),na.rm=T)
dem_dfc$demand_agric[dem_dfc$indicator=="total_demand"]<-sum(values(sdam_agric),na.rm=T)
dem_dfc$demand_pop[dem_dfc$indicator=="total_demand"]<-sum(values(sdam_pop),na.rm=T)
# total provided
dem_dfc$demand_artif[dem_dfc$indicator=="total_provided"]<-flow_df$demand_artif[flow_df$provision_cat=="contn_all"]
dem_dfc$demand_agric[dem_dfc$indicator=="total_provided"]<-flow_df$demand_agric[flow_df$provision_cat=="contn_all"]
dem_dfc$demand_pop[dem_dfc$indicator=="total_provided"]<-flow_df$demand_pop[flow_df$provision_cat=="contn_all"]
# unmet
dem_dfc$demand_artif[dem_dfc$indicator=="unmet_demand"]<-dem_dfc$demand_artif[dem_dfc$indicator=="total_demand"] - dem_dfc$demand_artif[dem_dfc$indicator=="total_provided"]
dem_dfc$demand_agric[dem_dfc$indicator=="unmet_demand"]<-dem_dfc$demand_agric[dem_dfc$indicator=="total_demand"] - dem_dfc$demand_agric[dem_dfc$indicator=="total_provided"]
dem_dfc$demand_pop[dem_dfc$indicator=="unmet_demand"]<-dem_dfc$demand_pop[dem_dfc$indicator=="total_demand"] - dem_dfc$demand_pop[dem_dfc$indicator=="total_provided"]
# relative unmet demand
dem_dfc$demand_artif[dem_dfc$indicator=="rel_unmet"]<-round(dem_dfc$demand_artif[dem_dfc$indicator=="unmet_demand"]/dem_dfc$demand_artif[dem_dfc$indicator=="total_demand"],digits=3)
dem_dfc$demand_agric[dem_dfc$indicator=="rel_unmet"]<-round(dem_dfc$demand_agric[dem_dfc$indicator=="unmet_demand"]/dem_dfc$demand_agric[dem_dfc$indicator=="total_demand"],digits=3)
dem_dfc$demand_pop[dem_dfc$indicator=="rel_unmet"]<-round(dem_dfc$demand_pop[dem_dfc$indicator=="unmet_demand"]/dem_dfc$demand_pop[dem_dfc$indicator=="total_demand"],digits=3)
# clean up
rm(sdam_agric,sdam_artif)
gc()
```
Here is the table
```{r}
# print table
kable(dem_dfc)
```


## 6.3 Plots by Catchment Area - binary SPA

The guidance note suggests to visualise results by catchment area. We do this here, providing plots of total service flow by catchment area. Total service flow is the sum of service flows to each cell within the area. Catchment areas for Norway are defined by NVE.

```{r, message = FALSE, results = 'hide', fig.height = 9}
# import shape
catch<-terra::vect(file.path(dfp,"Norway_files","NVE_vassdragomr","NVEKartdata",
                             "NVEData","Nedborfelt",
                             "Nedborfelt_Vassdragsomr.shp"))
# reproject
catch <- terra::project(catch,crs(lcm))
# remove svalbard
catch <- crop(catch,ext(4020798.64910517, 5135860.76845464, 3873563.87082421, 5500000))
# use terra::extract. This sums across cells within each polygon. A cell is
# within a polygon if its center is within the polygon.
# first we need a version where NA values are zeros, otherwise it returns NA everywhere
flow_d_econ_adj<-flow_d_econ
flow_d_econ_adj[is.na(flow_d_econ_adj)]<-0
sdam_econ_adj<-sdam_econ
sdam_econ_adj[is.na(sdam_econ_adj)]<-0
metr_d_econ <- flow_d_econ_adj/sdam_econ_adj
# get averages by area
metr_d_econ_by_catch <- terra::zonal(metr_d_econ,catch,fun = "mean",na.rm=T,touches=F)
# add data to the shapefile. We can do this because terra::zonal returns values in the same order as the attribute table of the spatial vector file
catch$metdecon <- metr_d_econ_by_catch[,1]
# plot
plot(catch,"metdecon",breaks=c(0,0.000000000001,0.2, 0.4, 0.6, 0.8, 1),col=viridis(6),main="Average proportion of demand (economic assets) met by supply")
## then the same for demand by population
flow_d_pop_adj<-flow_d_pop
flow_d_pop_adj[is.na(flow_d_pop_adj)]<-0
sdam_pop_adj<-sdam_pop
sdam_pop_adj[is.na(sdam_pop_adj)]<-0
metr_d_pop <- flow_d_pop_adj/sdam_pop_adj
# average by area
metr_d_pop_by_catch <- terra::zonal(metr_d_pop,catch,fun = "mean",na.rm=T,touches=F)
# add data to the shapefile. We can do this because terra::zonal returns values in the same order as the attribute table of the spatial vector file
catch$metdpop <- metr_d_pop_by_catch[,1]
# plot
plot(catch,"metdpop",breaks=c(0,0.000000000001,0.2, 0.4, 0.6, 0.8, 1),col=viridis(6), main="Average proportion of demand (population) met by supply")
# clean up
rm(catch,flow_d_econ_adj,sdam_econ_adj,flow_d_pop_adj,sdam_pop_adj,metr_d_econ_by_catch,metr_d_pop_by_catch,metr_d_econ,metr_d_pop)
gc()
```

## 6.4 Plots by Catchment Area - continuous SPA

The guidance note suggests to visualise results by catchment area. We do this here, providing plots of total service flow by catchment area. Total service flow is the sum of service flows to each cell within the area. Catchment areas for Norway are defined by NVE.

```{r, message = FALSE, results = 'hide', fig.height = 9}
# import shape
catch<-terra::vect(file.path(dfp,"Norway_files","NVE_vassdragomr","NVEKartdata",
                             "NVEData","Nedborfelt",
                             "Nedborfelt_Vassdragsomr.shp"))
# reproject
catch <- terra::project(catch,crs(lcm))
# remove svalbard
catch <- crop(catch,ext(4020798.64910517, 5135860.76845464, 3873563.87082421, 5500000))
# use terra::extract. This sums across cells within each polygon. A cell is
# within a polygon if its center is within the polygon.
# first we need a version where NA values are zeros, otherwise it returns NA everywhere
flow_d_econ_adj<-flow_d_econ_c
flow_d_econ_adj[is.na(flow_d_econ_adj)]<-0
sdam_econ_adj<-sdam_econ
sdam_econ_adj[is.na(sdam_econ_adj)]<-0
metr_d_econ <- flow_d_econ_adj/sdam_econ_adj
# get averages by area
metr_d_econ_by_catch <- terra::zonal(metr_d_econ,catch,fun = "mean",na.rm=T,touches=F)
# add data to the shapefile. We can do this because terra::zonal returns values in the same order as the attribute table of the spatial vector file
catch$metdecon <- metr_d_econ_by_catch[,1]
# plot
plot(catch,"metdecon",breaks=c(0,0.000000000001,0.2, 0.4, 0.6, 0.8, 1),col=viridis(6),main="Average proportion of demand (economic assets) met by supply")
## then the same for demand by population
flow_d_pop_adj<-flow_d_pop_c
flow_d_pop_adj[is.na(flow_d_pop_adj)]<-0
sdam_pop_adj<-sdam_pop
sdam_pop_adj[is.na(sdam_pop_adj)]<-0
metr_d_pop <- flow_d_pop_adj/sdam_pop_adj
# average by area
metr_d_pop_by_catch <- terra::zonal(metr_d_pop,catch,fun = "mean",na.rm=T,touches=F)
# add data to the shapefile. We can do this because terra::zonal returns values in the same order as the attribute table of the spatial vector file
catch$metdpop <- metr_d_pop_by_catch[,1]
# plot
plot(catch,"metdpop",breaks=c(0,0.000000000001,0.2, 0.4, 0.6, 0.8, 1),col=viridis(6), main="Average proportion of demand (population) met by supply")
# clean up
rm(catch,flow_d_econ_adj,sdam_econ_adj,flow_d_pop_adj,sdam_pop_adj,metr_d_econ_by_catch,metr_d_pop_by_catch,metr_d_econ,metr_d_pop)
gc()
```


# 7. Supply and Use Tables (SUT)

## 7.1 SUT Norway and NUTS2

### 7.1.1 Based on binary SPA

First we make a land cover map showing the supply categories we want to report in. And we need to aggregate our flows.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# so first make the vector of ecosystem types
estv<-c("Settlements and other artificial areas",
        "Cropland",
        "Grassland",
        "Forest and woodlands",
        "Heathlands and shrub",
        "Sparsely vegetated ecosystems",
        "Inland wetlands",
        "Rivers and canals",
        "Lakes and reservoirs",
        "Marine inlets and transitional waters",
        "Coastal beaches, dunes and wetlands",
        "Marine ecosystems")
# and add it to our lcm_idf
lcm_idf['es_type']<-NA
lcm_idf$es_type[lcm_idf$landcover_id %in% c(100,111,112,121,130,131,132,133,134,135,141,142,152)]<-estv[1] # settlem
lcm_idf$es_type[lcm_idf$landcover_id %in% c(200)]<-estv[2] # cropl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(310,320)]<-estv[3] # grassl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(400,410,420,440)]<-estv[4] # forest
lcm_idf$es_type[lcm_idf$landcover_id %in% c(520)]<-estv[5] # bush
lcm_idf$es_type[lcm_idf$landcover_id %in% c(600,611,623,631)]<-estv[6] # sparse
lcm_idf$es_type[lcm_idf$landcover_id %in% c(720)]<-estv[7] # wetlands
lcm_idf$es_type[lcm_idf$landcover_id %in% c(810)]<-estv[8] # rivers
lcm_idf$es_type[lcm_idf$landcover_id %in% c(910)]<-estv[9] # lakes
# category 10 doesn't exist
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1100)]<-estv[11] # coastal
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1200)]<-estv[12] # marine inlets
# make it a factor
#lcm_idf$es_type<-as.factor(lcm_idf$es_type)
#lcm_idf$es_type[lcm_idf$landcover_id %in% c(9900)]<-estv[12] # marine ecosystems
# Call it lcm5, which follows the required classification
lcm5 <- terra::subst(lcm2,lcm_idf$landcover_id,lcm_idf$es_type)
terra::plot(lcm5, main="Grunnkartet translated to SUT categories")
# clean up
rm(lcm2)
gc()
```

We want to disaggregate the flows by NUTS2 region. So for this we need the individual NUTS2 shapes. Then we compute the flow of service provision for each of the ecosystem categories in each region of interest. We do this by allocating the total service flow across our ecosystems of interest. Allocation follows weights, and the computation of those is specified in the guidance note. I make a function for this, which we can feed with any shapefile of interest for area-specific results.
```{r, message = FALSE,results='hide'}
# import n2
n2shp<-terra::vect(file.path(dfp,"Norway_files","NUTS2021_NO_LVL2_wosvalbard","NUTS2021_NO_LVL2_wos.shp"))
n2shp <- project(n2shp,crs(lcm))
# here is a function (written for easy debugging, not efficiency)
calc_es_flow_for_shp<-function(estv, # array of unique ecosystems
                               maskshp, # the area of interest as a spatial vector file
                               reflcm, # the reference land cover map
                               totalflowm, # the map of total service flows
                               refcnm # the map of reference curve number
){
  # Step 1: Crop and mask input data (if that is wanted)
  if(is.null(maskshp)){
    cnmc<-refcnm
    lcmc<-reflcm
    flowtc<-totalflowm
  } else {
    cnmc<-crop(refcnm,maskshp,mask=T)
    lcmc<-crop(reflcm,maskshp,mask=T)
    flowtc<-crop(totalflowm,maskshp,mask=T)
  }
  # Step 2: Make a vector of correction factors based on area-specific CN
  cfvc<-numeric()
  for(i in 1:length(estv)){
    cfvc[i]<-(100 - mean(cnmc[lcmc==estv[i]],na.rm=T))/100
  }
  # Step 3: Compute extent of each ecosystem
  exmc<-freq(lcmc)
  # Step 4: Move into one dataframe
  # then put all into one dataframe
  cfmc<-data.frame(cbind(estv,cfvc))
  dfwc<-cfmc %>% left_join(exmc,by=join_by(estv==value)) %>% 
    slice(order(factor(estv, levels = estv))) %>%
    mutate(cfvc = as.numeric(cfvc))
  # Step 5: Compute weighted extents
  dfwc['weighted_ext']<-dfwc$count*dfwc$cfvc
  sswc<-sum(dfwc$weighted_ext,na.rm=T) # sum weighted
  # Step 6: Compute allocation coefficients
  dfwc['alloc_coeff']<-dfwc$weighted_ext/sswc
  # Step 7: Compute total flow for the area of interest
  totalflowc<-sum(values(flowtc),na.rm=T)
  # Step 8: Compute disaggregated flows
  dfwc['fc_flow']<-dfwc$alloc_coeff*totalflowc
  # Step 9: Put all into one table
  dfwtc <- dfwc %>% select(c(estv,fc_flow)) %>% rename(ecosystem = estv) %>%
    add_row(ecosystem = "Total",fc_flow = totalflowc, .before = 1)
  # Step 10: Return result
  return(dfwtc)
}
```
So then we compute our flood protection service flows disaggregated by ecosystem and area. The unit is ha of economic assets protected from potential flood damage. Population is not considered here.
```{r, message = FALSE,results='hide'}
# Loop across our NUTS2 areas
sl<-list()
for(n in 1:length(unique(n2shp$NUTS_ID))){
  sl[[n]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = n2shp[n2shp$NUTS_ID == unique(n2shp$NUTS_ID)[n],],
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
  sl[[n]]['region'] <- unique(n2shp$NUTS_ID)[n]
}
# add for total country
ti<-length(sl)+1
sl[[ti]]<-calc_es_flow_for_shp(estv=estv,
                                         maskshp = NULL,
                                         reflcm = lcm5,
                                         totalflowm = flow_d_all_s_all,
                                         refcnm = cnm2[[4]])
sl[[ti]]['region'] <- "NO0"
# reformat to our final supply table
sdf<-do.call("rbind",sl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)
# clean up
rm(sl,ti)
gc()
```
Here is the supply table:
```{r}
# print table
kable(sdf)
```
And then the use table. For the time being we only report total use and we assume it goes to households.
```{r, message = FALSE,results='hide'}
udf <- sdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
# clean up
gc()
```
Here is the table:
```{r}
kable(udf)
```

### 7.1.2 Based on continuous SPA

So here we can use the same function, just feed it with the flows map based on continuous service provision.
```{r, message = FALSE,results='hide'}
# Loop across our NUTS2 areas
sl<-list()
for(n in 1:length(unique(n2shp$NUTS_ID))){
  sl[[n]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = n2shp[n2shp$NUTS_ID == unique(n2shp$NUTS_ID)[n],],
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
  sl[[n]]['region'] <- unique(n2shp$NUTS_ID)[n]
}
# add for total country
ti<-length(sl)+1
sl[[ti]]<-calc_es_flow_for_shp(estv=estv,
                                         maskshp = NULL,
                                         reflcm = lcm5,
                                         totalflowm = flow_d_econ_c,
                                         refcnm = cnm2[[4]])
sl[[ti]]['region'] <- "NO0"
# reformat to our final supply table
sdf<-do.call("rbind",sl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)
# clean up
rm(sl,ti)
gc()
```
Here is the supply table:
```{r}
# print table
kable(sdf)
```
And then the use table. For the time being we only report total use and we assume it goes to households.
```{r, message = FALSE,results='hide'}
udf <- sdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
# clean up
gc()
```
Here is the table:
```{r}
kable(udf)
```


## 7.2 SUT KS Project

### 7.2.1 Based on binary SPA

And here for our case study municipalities
```{r, message = FALSE,results='hide'}
# Loop across our NUTS2 areas
kl<-list()
kl[[1]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ahs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[1]]['region'] <- "Aurskog-Holand" 
kl[[2]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ehs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[2]]['region'] <- "Evje og Hornnes" 
kl[[3]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = hos,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[3]]['region'] <- "Holtaalen" 
kl[[4]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = mos,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[4]]['region'] <- "Moss" 
kl[[5]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = nrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[5]]['region'] <- "Nordreisa" 
kl[[6]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = sds,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[6]]['region'] <- "Stad" 
kl[[7]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[7]]['region'] <- "Vardo" 
kl[[8]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vas,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[8]]['region'] <- "Vaagan" 
# reformat to our final supply table
kdf<-do.call("rbind",kl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)
# create use table
kudf <- kdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
#clean up
gc()
```
Here is the supply table:
```{r}
# print table
kable(kdf)
```

And here is the use table:
```{r}
kable(kudf)
```

### 7.2.2 Based on continuous SPA

And here for our case study municipalities
```{r, message = FALSE,results='hide'}
# Loop across our NUTS2 areas
kl<-list()
kl[[1]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ahs,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[1]]['region'] <- "Aurskog-Holand" 
kl[[2]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ehs,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[2]]['region'] <- "Evje og Hornnes" 
kl[[3]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = hos,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[3]]['region'] <- "Holtaalen" 
kl[[4]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = mos,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[4]]['region'] <- "Moss" 
kl[[5]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = nrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[5]]['region'] <- "Nordreisa" 
kl[[6]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = sds,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[6]]['region'] <- "Stad" 
kl[[7]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[7]]['region'] <- "Vardo" 
kl[[8]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vas,
                                reflcm = lcm5,
                                totalflowm = flow_d_econ_c,
                                refcnm = cnm2[[4]])
kl[[8]]['region'] <- "Vaagan" 
# reformat to our final supply table
kdf<-do.call("rbind",kl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)
# create use table
kudf <- kdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
#clean up
rm(lcm5)
gc()
```
Here is the supply table:
```{r}
# print table
kable(kdf)
```

And here is the use table:
```{r}
kable(kudf)
```


# 8. Ideas for Improvements

-   Add street infrastructure from OSM to our artificial areas - a modification to consider if the next version of Grunnkartet does not capture infrastructure well
-   Feed in finer DEM data for more accurate modeling of water flows. 
-   Add an instructive example of how service provision is modeled, using a hypothetical 20x20 grid