---
title: "Ecosystem Service Flood Control | Norway"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cosmo
    code_folding: hide
---

```{css, echo=FALSE}
table {
  float: left;
}
```

Magnus Merkle (SSB), with Marte Lindseth (SSB), Erik Stange (NINA), and Trine Randen (SSB) 

Naturregnskapsprosjekt, 5. Sepetmber 2025

\

# Changelog

Changes since version 1:

-   Added a statement on limitations to the first section
-   Added all KS municipalities as case studies
-   Some restructuring of sections for easier oversight
-   Reduced number of plots in sections 2 - 4
-   Added a table of data sources

# Abstract and Limitation

In this notebook we quantify the ecosystem service of flood control for the Norwegian mainland. Most operations are executed with functions from R packages dplyr and terra. The notebook is created on the basis of the following sources:

-   Vallecillo et al. (2019) Ecosystem services accounting Part II Pilot accounts for crop and timber provision, global climate regulation and flood control. European Commission: JRC Technical Reports.
-   EC (2023) Guidance note to account for flood control ecosystem service - 1st Draft. Agenda Item 6, Task Force on Ecosystem Accuonting, 28-29 November 2023.
-   Immerzeel & Korkou (2024) Evaluation of the INCA-Tool for the purpose of ecosystem services accounting in Norway. EU Grants Project 2022-NO-EGD- Work Package 2. Norwegian Institute for Nature Research

The quantification process is implemented with reference to the work done by Harry Hodgins, Statistics Ireland. Harry has quantified ES flood control for Ireland using R. We do not copy the Statitics Ireland script in every aspect, but we use it as inspiration for the approach implemented here.

The quantification process follows the EC guidance note. It is important to note that the objective of the process is to identify the contribution of different ecosystems to the provision of flood control (i.e. reduced flood risk) and how this provision changes over time. The process is not designed to provide accurate predictions of the spatial pattern of flood risk and flood risk mitigation at fine spatial resolution. It is not a dynamic hydrological model, but rather an accounting exercise. Therefore, the resulting maps and tables should be interpreted with care and with the limitations of the process in mind.

# 1. Filepaths and Functions

The main functions we use are part of dplyr and terra.
```{r, results = 'hide', message = FALSE, warning=FALSE}
# make space
rm(list = ls())
gc()
# packages
library(dplyr)
library(tidyr)
library(terra)
library(sf)
library(osmdata)
library(knitr)
# filepath
dfp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "data","INCA")
# A function to extract levels vectors from factor data stored in rasters 
# rst is raster file, lrn is layer name
extr_lvl_from_raster_lyr<-function(rst,lrn){
  if(is.factor(rst[[lrn]])){
    v1 <- as.vector(values(rst[[lrn]]))
    v2 <- as.vector(levels(rst[[lrn]]))[[1]]
    v3 <- v2$value[v1]
    return(v3)
  } else {
    return(as.vector(values(rst[[lrn]])))
  }
}
# A function to download roads data for specific shapes within Norway
osm_get_streets_for_KS<-function(kshp,lcm){
  # crop and reproject lcm
  k_lcm<-crop(lcm,kshp,mask=T)
  k_lcm_w84 <- project(k_lcm, "EPSG:4326")
  # download roads
  roads <- k_lcm_w84 %>% 
    sf::st_bbox() %>% 
    osmdata::opq() %>%
    osmdata::add_osm_feature(key="highway",value = c("primary", "secondary", "tertiary", "trunk", "motorway")) %>% 
    osmdata_sf()
  # reproject
  roads_t <- roads$osm_lines %>% 
    terra::vect() %>%
    terra::project(lcm)
  # return result
  return(roads_t)
}
```

# 2. Data Sources

Here is a table of data sources used in this project.
```{r}

dfs<-data.frame(matrix(nrow=10,ncol=8))
colnames(dfs) <- c("Input to","Data","Name","Provider","Resolution","Year","Future Updates","Notes")

dfs[1,]<-c("SPA and SDA",
           "Reference land cover map (LCM) with lookup table",
           "Grunnkart_OT",
           "Kartverket, NIBIO, SSB, Mdir",
           "100m grid",
           "2024",
           "2025",
           "This is probably the most important input because we derive ecosystems, curve numbers, and economic assets from it. Updates should ideally be used to quantify changes in service flows over time. Comparability over time would be relevant for this purpose.")
dfs[2,]<-c("SPA",
           "Hydrological soil type map",
           "hydro_soilgroup_ INCA_EPSG3035",
           "Eurostat (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "Unknown",
           "Unknown",
           "This map is rather coarse, containing only four categories. A part of it in Northern Norway is missing. It would probably be good to replace this with something more accurate/specific for Norway.")
dfs[3,]<-c("SPA",
           "Curve Number Lookup Table",
           "curve_number_per_ landcover_soil",
           "Eurostat (INCA default version_2_2)",
           "Not spatial",
           "Probably not relevant",
           "Not required",
           "Some Norwegian land cover categories are not included in the current table. For example: glaciers. It would be good to update the table so it contains all land cover categories that exist in Norway. Alternatively it would be good with a curve number lookup table for all categories reported in the Norwegian LCM. Then we could avoid the translation into CORINE categories.")
dfs[4,]<-c("SPA",
           "Impervious Density Map",
           "IMD_2018_100m_ eu_03035_V2_0_INCA_ 100m_EPSG3035",
           "Eurostat (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "2018",
           "Unknown",
           "Here as well it might be possible to get hold of better data for the Norwegian context.")
dfs[5,]<-c("SPA and service flow",
           "Slope Map",
           "dt1_slope_INCA_ EPSG3035_ 100m_Byte",
           "European Space Agency (INCA default version_2_2)",
           "100m grid (rescaled)",
           "Probably not relevant",
           "Probably not relevant",
           "We could consider trying a finer resolution slope map in the flow quantification module. To test whether this might make results more accurate. Either using Norwegian elevation data, or by importing the original copernicus data.")
dfs[6,]<-c("SPA",
           "Riparian Zones Map",
           "RiparianZonesEU_ INCA_EPSG3035",
           "EC JRC (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "Unknown",
           "Unknown",
           "If NVE has more accurate data for riparian zones we could try replacing this input. Also: if such zones can change over time, it would be relevant to update this data input over time.")
dfs[7,]<-c("SDA",
           "Flood Plains Map",
           "floodMapEU_rp500y",
           "Eurostat (INCA default version_2_2)",
           "100m grid (rescaled?)",
           "Probably not relevant",
           "Probably not relevant",
           "We currently use data for 1 in 500yrs floods. We could change this to 1 in 100 yrs (then service flows will be lower) or 1 in 1000 yrs (then service flows will be higher)")
dfs[8,]<-c("SDA",
           "Population Map",
           "GHS_POP_E2020_ GLOBE_R2023A_ 54009_100_V1_0",
           "GHSL (INCA default version_2_2)",
           "100m grid (rescaled)",
           "2020",
           "Yes (every 5 years)",
           "This is another key input which should be varied over time to quantify changes in service flow.")
dfs[9,]<-c("Service Flow",
           "Elevation Map",
           "dt1_dem_INCA_ EPSG3035_100m",
           "European Space Agency (INCA default version_2_2)",
           "100m grid (rescaled)",
           "Probably not relevant",
           "Probably notrelevant",
           "We could try a finer resolution map to see whether that changes results. We could also include areas outside Norway if cross-border flows are flood-relevant for Norway.")
dfs[10,]<-c("Service Flow",
           "Various shapefiles",
           "Various",
           "Various",
           "Vectors",
           "Not relevant",
           "Not relevant",
           "Only to retrieve results for specific subnational regions.")

kable(dfs)
```

# 3. Service Provision (SPA)

## 3.1 Overview

The first step in the quantification of flood control is to create a raster map of potential runoff retention. This map is a key input to the ensuing steps. We start from the Norwegian land cover map and a European Hydrological Soil Charateristics map. Each combination of land cover and soil characteristic is associated to a Curve Number (CN) value - an index of infiltration/runoff. The resulting curve number raster map is then further refined using various inputs. Once the runoff retention map is obtained, we derive Service Providing Areas (SPA) from it using specific thresholds.

## 3.2 Reference CN Map

### 3.2.1 LCM

Read in and plot our land cover categories.
```{r, fig.height = 9}
# read Norway Grunnkart (this is our reference LCM) ----------------------------
lcm<-terra::rast(file.path(dfp,"Norway_files","Grunnkart","Grunnkart_OT.tif"))
unique(lcm)
terra::plot(lcm, main="Norwegian Land Cover Map (Grunnkart)")
```

Now the fact of the matter is that the categories used in the Norwegian LCM do not perfectly correspond to the categories used in the curve number table. Therefore we first need to translate the Norwegian categories into the system used in the CORINE data.
```{r,message = FALSE, results = 'hide', fig.height = 9}
# make a numeric version of the lcm --------------------------------------------
lcm2<-lcm
activeCat(lcm2) <- "Value"
lcm2<-as.numeric(lcm2)


# read in grunnkart lookup -----------------------------------------------------
lcm_idf<-read.csv(file.path(dfp,"Norway_files","Grunnkart",
                            "Lookup_grunnkart.csv"))

# read in curve number table ---------------------------------------------------
cn_idf<-read.csv(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                           "Lookup_tables",
                           "curve_number_per_landcover_soil.csv"))

# adjust lcm to feature official CORINE categories -----------------------------
# (otherwise our CN lookup table does not fit)
# I create this concordance using the following sources
# https://land.copernicus.eu/content/corine-land-cover-nomenclature-guidelines/html/
# https://nibio.brage.unit.no/nibio-xmlui/bitstream/handle/11250/3120510/NIBIO_RAPPORT_2024_10_28.pdf?sequence=1&isAllowed=y
lcm_idf['CN_cover_id']<-NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id %in% 
                      c(100,111,112,121,130,131,132,133,134,135,142,152)] <- 1
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 141] <- 141
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 200] <- 241
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 310] <- 231
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 320] <- 321
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 400] <- 312
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 410] <- 311
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 420] <- 312
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 440] <- 313
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 520] <- 322
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 600] <- 333
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 611] <- 332
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 623] <- 333
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 631] <- NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 720] <- 322
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 810] <- NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 910] <- NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 1100] <- NA
lcm_idf$CN_cover_id[lcm_idf$landcover_id == 1200] <- NA

# and then we add the last category that is not available (uklassifisert)
lcm_idf <- rbind(lcm_idf,c(NA,9900,"unclassified","uklassifisert",NA))
lcm_idf$landcover_id <- as.numeric(lcm_idf$landcover_id)

# make a new lcm that features the CN_cover_id
lcm3 <- terra::subst(lcm2,lcm_idf$landcover_id,lcm_idf$CN_cover_id)
terra::plot(lcm3, main="Grunnkartet translated to CORINE LCM categories")

# make a numeric version of this
lcm4 <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
values(lcm4)<-as.numeric(extr_lvl_from_raster_lyr(lcm3,"value"))
```

So this is our reference land cover map. Moving on.

### 3.2.2 Hydrological Soil Type

```{r}
# read hydrological soil type map ----------------------------------------------
soils<-terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "Hydrological_soil_type_map",
                             "hydro_soilgroup_INCA_EPSG3035.tif"))
# mm comment: only four categories seems a bit coarse...
# mm comment: the soils data is masked in a weird way in the north of Norway. 
# I think we should really use something better here.
```
Ok so a few adjustments are required here. First we need to mask it to our reference LCM. Then we need to figure out what we do with the part in northern Norway that is missing in the data.
```{r,message = FALSE, results = 'hide', fig.height = 9}
# first crop and mask soils ----------------------------------------------------
soils<-crop(soils,lcm)
soils<-mask(soils,lcm)
terra::plot(soils, main="Hydrological soil type masked to Norway")
# mm comment: so here you see again how the soil map is weirdly masked in the 
# north. comes from the input data. To be sorted out later
```

Here is my suggested quick fix (which should be revisited later): We just define the missing data cells as hydrological soil type 2. This is the most common hydrological soil type in Norway.
```{r,message = FALSE, results = 'hide', fig.height = 9}
soils[is.na(soils)]<-2
soils<-mask(soils,lcm3)
terra::plot(soils, main="Adjusted hydrological soil type for Norway")
```

Ok this is better. The map has a quite a few holes, and this comes from cells that have no CN number in tghe lookup table. Revisit later. The categories in question are: 9900 (uklassifisert), 1200 (marine økosystemer (uspesifisert)), 1100 (Kyststrender dyner og våtmark (uspesifisert)), 910 (Innsjøer og dammer), 810 (Elver og bekker), and 631 (Isbreer og varige snøfonner). We might want to consider making a modification later so that we assume some CN for some of these categories. Could for example check how Switzerland treats their glacier land cover. Assuming no CN implies that these cells cannot provide flood control.

### 2.2.3 Curve Number

So this is now basically a lookup exercise. My approach is to vectorise the rasters, then run the merge process, then rasterise again. There are probably more efficient ways to do this, but this one is simple and it seems to work.
```{r,message = FALSE, results = 'hide', fig.height = 9}
# extract data vectors
lcmv <-extr_lvl_from_raster_lyr(lcm3,"value")
soilsv <-extr_lvl_from_raster_lyr(soils,"hydro_soilgroup_INCA_EPSG3035")
# put into a dataframe
dfc<-as.data.frame(matrix(NA,nrow=length(lcmv),ncol=2))
colnames(dfc)<-c("lcmd","soilsd")
dfc$lcmd <- as.numeric(lcmv)
dfc$soilsd <- soilsv
# join with CN lookup table
dfc <- dfc %>% left_join(cn_idf,by = join_by(lcmd == landcover_id,soilsd == Soil))
# then make a raster from the vector
cnm <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
values(cnm)<-dfc$CN
# plot result
terra::plot(cnm, range=c(0,100), main="Norway Curve Number Map", sub="High number implies high runoff, low number implies low runoff")
```
```{r, message = FALSE, results = 'hide'}
# clean up ---------------------------------------------------------------------
rm(dfc,soils,lcmv,soilsv)
gc()
```

## 3.3 CN Map Modification

### 3.3.1 Imperviousness Correction

We use the INCA default impervious density map.
```{r}
imp <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "Impervious_density_map",
                             "IMD_2018_100m_eu_03035_V2_0_INCA_100m_EPSG3035.tif"))
```
Mask to our reference raster.
```{r,message = FALSE, results = 'hide'}
# first crop and mask to Norway
imp<-crop(imp,lcm)
imp<-mask(imp,lcm)
terra::plot(imp, main="Impervious Density Norway")
```

For this operation we do not need to vectorise first. The equation comes from the guidance note, corrected by Harry. Can just do this straight with the rasters. And then plot the result.
```{r, message = FALSE, results = 'hide', fig.width = 12}
cnmi <- 98*(imp/100) + (1-(imp/100))*cnm
# make a stack
cnm1 <- c(cnm,cnmi)
names(cnm1)<-c("CN","CN Adjusted 1")
```
Differences are visible mostly in cities. Not a big difference. Here zooming in on Oslo.
```{r, message = FALSE, results = 'hide', fig.width = 12}
terra::plot(cnm1,range=c(0,100),ext = c(4200000,4400000,4000000,4200000))
```
```{r, message = FALSE, results = 'hide'}
# clean up
rm(imp,cnmi)
gc()
```

### 3.3.2 Slope Correction

We use the INCA default DEM and Slope maps, cropped and masked to Norway.
```{r,message = FALSE, results = 'hide', fig.height = 9}
slop <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "DEM_and_Slope_maps",
                             "dt1_slope_INCA_EPSG3035_100m_Byte.tif"))
slop<-crop(slop,lcm)
slop<-mask(slop,lcm)
plot(slop, main="Slope Norway [expressed in %")
```

Our data provides slopes in percent. The equation stated in the guidance note based on Huang et al 2016 requires slopes as ratio (ratio = percent/100). Zero slope is NA in the data, so I first turn all NA to zero before running the process. Then apply the formula as stated in the guidance note. And then plot to see what difference the operation makes.
```{r, message = FALSE, results = 'hide', fig.width = 12}
slop[is.na(slop)]<-0
cnma <- cnm1[['CN Adjusted 1']] * ((322.79 + (15.63*slop/100))/((slop/100)+323.52))
cnm2<-c(cnm1,cnma)
names(cnm2)<-c("CN","CN Adjusted 1","CN Adjusted 1_2")
```

The difference is really small. We need to zoom in quite a lot to see anything. Here for example, somewhere close to Åndalsnes.
```{r, message = FALSE, results = 'hide', fig.width = 12}
plot(cnm2[[c(2,3)]],range=c(0,100),ext = c(4200000,4210000,4370000,4380000))
```
```{r, message = FALSE, results = 'hide'}
rm(slop,cnma,cnm1)
gc()
```

### 3.3.3 Riparian Zone Correction

Then the riparian zones. Again we use INCA default data for this adjustment.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# import data
rip <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                              "Riparian_zones_map",
                              "RiparianZonesEU_INCA_EPSG3035.tif"))


# first crop and mask to Norway
rip<-crop(rip,lcm)
rip<-mask(rip,lcm)
# to make it better visible
rip[is.na(rip) & !is.na(lcm3)]<-0

plot(rip,main="Riparian Zones Norway")
```

Ok and then the process.
```{r, message = FALSE, results = 'hide', fig.width = 12}

# define relevant landcover categories as per guidance note
lcm_idf['rip_relevant']<-0

relevant_codes <- c(244,311,312,313,321,322,323,324,411,412,421,422,423)
# mm note: most of those do not exist in our Grunnkart in Norway...
# the only ones in Norway are category 3 (forest)
lcm_idf$rip_relevant[lcm_idf$CN_cover_id %in% relevant_codes]<-1

# now for those cells where we are in a relevant riparian zone we will attribute
# the lowest possible cn number for the respective land cover. 

# get vector of rip
ripv <- extr_lvl_from_raster_lyr(rip,"RiparianZonesEU_INCA_EPSG3035")
# get vector of lcm
lcmv <- extr_lvl_from_raster_lyr(lcm3,"value")
lcmv <-as.numeric(lcmv)

# get vector of CN
cnv <- extr_lvl_from_raster_lyr(cnm2,"CN Adjusted 1_2")

# relevant rip lcmv
dfc<-as.data.frame(matrix(NA,ncol=3,nrow=length(lcmv)))
colnames(dfc)<-c("lcm","rip","relevant")

#first the land cover data
dfc$lcm <- lcmv
#then the riparian zone data
dfc$rip <- ripv
# then the relevant cells
dfc$relevant[dfc$rip==1 & dfc$lcm %in% relevant_codes] <- 1
# then find the lowest possible CN value for each cell
dfc <- dfc %>% 
  left_join(cn_idf %>% filter(Soil==1) %>% select(-Soil),by = join_by(lcm == landcover_id)) %>%
  rename(CN1 = CN)
#clean up
rm(rip,ripv,lcmv)
dfc <- dfc %>% select(-c(lcm,rip))
gc()
#then get the cn number for each cell
dfc['CNIS'] <- cnv
rm(cnv)
#adjusted CN is the lowest CN
dfc['CNISR'] <- dfc$CN1
gc()
# apart from in rip-irrelevant cells, where it is the old CN value
dfc$CNISR[is.na(dfc$relevant)] <- dfc$CNIS[is.na(dfc$relevant)]
gc()

# then make a new raster
cnma <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
values(cnma)<-dfc$CNISR
names(cnma)<-c("CN Adjusted 1_2_3")

# then stack with our other 
cnm2 <- c(cnm2,cnma)

# clean up
rm(dfc,relevant_codes,cnma)
gc()
```
Again we don't see so much on the large picture. Let's zoom in on where Svelle (Øyeren/Glomma) flows into Nitelva (passing Lillestrøm).
```{r, message = FALSE, results = 'hide', fig.width = 12}
plot(cnm2[[c(3,4)]],range=c(0,100),ext = c(4380000,4384000,4091000,4095000))
```

That was it for adjustments. The exercise might be much more precise if we could get hold of more detailed/specific data for Norway.

## 2.4 Runoff Retention Map

We just need to reverse the scale for this. So that a high number implies that water is absorbed, and a low number implies that water is not absorbed. This is now the main map from which we derive flood control service provision.
```{r, message = FALSE, results = 'hide', fig.height = 9}

max_value <- as.numeric(global(cnm2[[4]],fun = "max", na.rm=T))

potm <- max_value - cnm2[[4]]
names(potm)<-"Pot_Runuff_Retention"

plot(potm, range=c(0,100),main = "Potential Runoff Retention Map")

rm(max_value)
```

## 3.5 SPA Thresholds

So now based on our potential runoff retention map we define our service providing areas (SPA). There are three different ones, depending on whether the areas are located on urban/artificial land, agricultural land, or natural land. SP areas are binary, so 1/0. In order to define them, we use runoff retention thresholds, one for each parent land category. Harry at Statistics Ireland came up with their own thresholds, calculated the same way as the EU guidance note suggests. For our version I have first tried to use the thresholds that the guidance note suggests for Europe (27 for artificial land, 52 for agricultural land, 61 for natural land), but then found that these may not be so applicable for Norway. One problem was, for example, that none of the Norwegian agricultural land could be classified as flood control service providing area, simply because the runoff retention threshold of 52 is too high (there is no agricultural patch in Norway that exceeds this threshold).

I therefore calculate Norway specific thresholds using the approach that the guidance note suggests. However, I think this should be revisited. Here are the thresholds
```{r, message = FALSE, results = 'hide'}

## first on artificial
artif_id<-c(1,141)
spam_artif_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_artif_am[[1]]<-0
spam_artif_am[is.na(lcm3)]<-NA
for(i in 1:length(artif_id)){
  spam_artif_am[lcm4 == artif_id[i]]<-mean(potm[lcm4 == artif_id[i]],na.rm=T) + sd(potm[lcm4 == artif_id[i]],na.rm=T)
}
spa_tld_artif <- mean(spam_artif_am[lcm4 %in% artif_id],na.rm=T)

## then on agricultural
agric_id<-c(241,231)
spam_agric_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_agric_am[[1]]<-0
spam_agric_am[is.na(lcm3)]<-NA
for(i in 1:length(agric_id)){
  spam_agric_am[lcm4 == agric_id[i]]<-mean(potm[lcm4 == agric_id[i]],na.rm=T) + sd(potm[lcm4 == agric_id[i]],na.rm=T)
}
spa_tld_agric <- mean(spam_agric_am[lcm4 %in% agric_id],na.rm=T)

## then on natural
natur_id<-c(321,312,311,313,322,333,332)
spam_natur_am <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_natur_am[[1]]<-0
spam_natur_am[is.na(lcm3)]<-NA
for(i in 1:length(natur_id)){
  spam_natur_am[lcm4 == natur_id[i]]<-mean(potm[lcm4 == natur_id[i]],na.rm=T) - sd(potm[lcm4 == natur_id[i]],na.rm=T)
}
spa_tld_natur <- mean(spam_natur_am[lcm4 %in% natur_id],na.rm=T)

## clean up
rm(spam_artif_am,spam_agric_am,spam_natur_am)
gc()
```
```{r}
print(paste("Artificial tld:",spa_tld_artif))
print(paste("Agricultural tld:",spa_tld_agric))
print(paste("Natural tld:",spa_tld_natur))
```

## 3.6 SPA Maps

Now given our thresholds we make a stack of maps for the SPAs
```{r, message = FALSE, results = 'hide', fig.height = 9}
gc()
spam_artif<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_artif[[1]]<-0
spam_artif[is.na(lcm3)]<-NA
spam_artif[lcm4 %in% artif_id & potm >= spa_tld_artif] <-1

spam_agric<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_agric[[1]]<-0
spam_agric[is.na(lcm3)]<-NA
spam_agric[lcm4 %in% agric_id & potm >= spa_tld_agric] <-1

spam_natur<-rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
spam_natur[[1]]<-0
spam_natur[is.na(lcm3)]<-NA
spam_natur[lcm4 %in% natur_id & potm >= spa_tld_natur] <-1

spam <-c(spam_artif,spam_agric,spam_natur)
names(spam)<-c("SPA Artificial","SPA Agricultural", "SPA Natural")
ccspam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(spam[[1]],col=ccspam,colNA="grey40",main = "SPA Artificial")
plot(spam[[2]],col=ccspam,colNA="grey40",main = "SPA Agricultural")
plot(spam[[3]],col=ccspam,colNA="grey40",main = "SPA Natural")
rm(ccspam,spam_artif,spam_agric,spam_natur)
```

# 4. Service Demanding Areas (SDA)

## 4.1 Calculation and Results

SDAs are service demanding areas. These are defined as economic assets located in floodplains. Formally: total spatial extent of economic assets that could be potentiall affected by a 1 in 500 years flood. Can also run this on the basis of a 1 in 100 years flood. Harry (Statistics Ireland) used 1 in a 1000 years flood. We identify SDAs seperately for agricultural land and for artificial land. Second possibility: total amount of population inhabiting in SDA (Harry does not do this, but we can). So first, we start reading in the data and making it fit to our reference map.
```{r, message = FALSE, results = 'hide', fig.height = 9}
# population
popr<-rast(file.path(dfp,"population","Mollweide","GHS_POP_E2020_GLOBE_R2023A_54009_100_V1_0.tif"))
popr<-crop(popr,ext(250000,
                    2000000,
                    6600000,
                    8000000) # 8700000 if svalbard included
                    )
popr <- terra::project(popr,lcm,mask=T)

# flood
flood <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                             "Flood_plains",
                             "floodMapEU_rp500y.tif"))
flood <- project(flood,crs(lcm))
flood<-crop(flood,lcm)
ext(flood)<-align(ext(flood),lcm)
flood<-mask(flood,lcm)
plot(flood, main = "Flood Plains Norway")
```

Now it is literally just a question of identifying cells in the flood areas. Different levels in the flood map relate to different depths of flood. Let's just define any number \>0 as flood. Could later consider to make it more continuous.
```{r, message = FALSE, results = 'hide'}
sdam_pop <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_pop[[1]] <- 0
sdam_pop[is.na(lcm3)]<-NA
sdam_pop[flood >0 & popr>0] <- popr[flood >0 & popr>0]

sdam_artif <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_artif[[1]] <- 0
sdam_artif[is.na(lcm3)]<-NA
sdam_artif[flood >0 & lcm4 %in% artif_id] <- 1

sdam_agric <- rast(ext(lcm3),resolution=res(lcm3),crs=crs(lcm3))
sdam_agric[[1]] <- 0
sdam_agric[is.na(lcm3)]<-NA
sdam_agric[flood >0 & lcm4 %in% agric_id] <- 1

sdam_econ <- sdam_artif + sdam_agric
```
Then let's plot the results here.
```{r, message = FALSE, results = 'hide', fig.height = 9}
plot(sdam_agric,col=c("grey90","blue"),main = "Agricultural land in floodplains")
plot(sdam_artif,col=c("grey90","blue"),main = "Artificial land in floodplains")
ccsdam<-data.frame(from=c(0,0.00001),to=c(0,Inf),col=c("grey90","blue"))
plot(sdam_pop,col=ccsdam, main = "Population in floodplains")

rm(popr,flood,ccsdam)
gc()
```

Important to note for the next sections: "Economic assets" are defined as cells that are either agricultural land or artificial land. So we add two service demand categories together for this.

## 4.2 Plots for KS Project

First we need to import all KS shapefiles.
```{r, message = FALSE, results = 'hide'}
ahs<-terra::vect(file.path(dfp,"Norway_files","Aurskog","aurskog.shp"))
ehs<-terra::vect(file.path(dfp,"Norway_files","Evje","evje_formatted.shp"))
hos<-terra::vect(file.path(dfp,"Norway_files","Holtalen","holtalen.shp"))
mos<-terra::vect(file.path(dfp,"Norway_files","Moss","moss.shp"))
nrs<-terra::vect(file.path(dfp,"Norway_files","Nordreisa","nordreisa.shp"))
sds<-terra::vect(file.path(dfp,"Norway_files","Stad","stad.shp"))
vrs<-terra::vect(file.path(dfp,"Norway_files","Vardo","vardo.shp"))
vas<-terra::vect(file.path(dfp,"Norway_files","Vaagan","vaagan.shp"))
```
And then let's plot service demand for our case studies.
```{r, message = FALSE, results = 'hide', fig.height = 7}
gc()
# Aurskog Høland
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(ahs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Aurskog Holand") 
stv<-osm_get_streets_for_KS(ahs,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(ahs,mask=T),colNA="grey40",
     main="Population in floodplains, Aurskog Holand") 
stv<-osm_get_streets_for_KS(ahs,lcm)
plot(stv,add=T)
# Evje and Hornnes
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(ehs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Evje and Hornnes") 
stv<-osm_get_streets_for_KS(ehs,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(ehs,mask=T),colNA="grey40",
     main="Population in floodplains, Evje and Hornnes") 
stv<-osm_get_streets_for_KS(ehs,lcm)
plot(stv,add=T)
# Holtålen
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(hos,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Holtaalen") 
stv<-osm_get_streets_for_KS(hos,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(hos,mask=T),colNA="grey40",
     main="Population in floodplains, Holtaalen") 
stv<-osm_get_streets_for_KS(hos,lcm)
plot(stv,add=T)
#Moss
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(mos,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Moss") 
stv<-osm_get_streets_for_KS(mos,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(mos,mask=T),colNA="grey40",
     main="Population in floodplains, Moss") 
stv<-osm_get_streets_for_KS(mos,lcm)
plot(stv,add=T)
# Nordreisa
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(nrs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Nordreisa") 
stv<-osm_get_streets_for_KS(nrs,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(nrs,mask=T),colNA="grey40",
     main="Population in floodplains, Nordreisa") 
stv<-osm_get_streets_for_KS(nrs,lcm)
plot(stv,add=T)
# Stad
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(sds,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Stad") 
stv<-osm_get_streets_for_KS(sds,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(sds,mask=T),colNA="grey40",
     main="Population in floodplains, Stad") 
stv<-osm_get_streets_for_KS(sds,lcm)
plot(stv,add=T)
# Vardø
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(vrs,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Vardo") 
stv<-osm_get_streets_for_KS(vrs,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(vrs,mask=T),colNA="grey40",
     main="Population in floodplains, Vardo") 
stv<-osm_get_streets_for_KS(vrs,lcm)
plot(stv,add=T)
# Vågan
ccsdam<-data.frame(value=c(0,1),col=c("grey90","blue"))
plot(sdam_econ %>% crop(vas,mask=T),col=ccsdam,colNA="grey40",
     main="Economic assets in floodplains, Vaagan") 
stv<-osm_get_streets_for_KS(vas,lcm)
plot(stv,add=T)
plot(sdam_pop %>% crop(vas,mask=T),colNA="grey40",
     main="Population in floodplains, Vaagan") 
stv<-osm_get_streets_for_KS(vas,lcm)
plot(stv,add=T)
```

# 5. Service Flow SPA --> SDA

## 5.1 Introduction

Now the service flow is quantified at the level of the SDAs. We calculate it for each SDA gridcell as the ratio of upstream surface area that is covered by SPA. Upstream surface area is the area from which we expect water to flow - so this requires slope information.

NB: the guidance suggests to use all upstream area for this, irrespective of political boundaries. For the time being we limit ourselves to Norway as service providing area, but could include neighbouring countries as well if we want.

The calculation of service flow is straightforward using the built-in functions of R's terra package. First, we compute the number of upstream cells for each SDA cell. Second, we compute the proportion of these upstream cells that are covered by SPA (of various kind). The service flow is then equal to the ratio of upstream cells covered by SPA over all upstream cells.

Example: If the the ratio is 1, then all upstream cells of a service demanding cell are covered by SPA, hence "flood controlled by the ecosystem". If the ratio is 0, then none of the upstream cells of a service demanding cell are covered b y SPA, hence "not flood controlled by the ecosystem".

Important to note here is that when there is no unique flow direction to identify from the slope data, then terra will allocate the flow direction randomly. This is not ideal, but at least we can make terra return unique results by setting seeds for the random sequence generator.

## 5.2 Flow Accumulation

First we import elevation data and compute flow direction and flow accumulation for each grid cell.
```{r, message = FALSE, results = 'hide', fig.height = 9}

elev <- terra::rast(file.path(dfp,"version_2_2","inca_input_floodcontrol",
                              "DEM_and_Slope_maps",
                              "dt1_dem_INCA_EPSG3035_100m.tif"))
elev<-crop(elev,lcm)
elev<-mask(elev,lcm)
plot(elev,main="Norway Elevation")

set.seed(2025) # I do this for replicability. Flowdir allocates direction randomly when it is impossible to find a unique direction from the elevation data
flowdir_total <- terra::terrain(elev,v="flowdir")

flow_acc_total <- terra::flowAccumulation((flowdir_total))
plot(flow_acc_total, main = "Flow accumulation (number of cells upstream flowing into each cell",col=data.frame(from=c(0,2,10,100,500,1000),
                    to=c(2,10,100,500,1000,8000),
                    col=c("grey90","steelblue1","steelblue4","lightsalmon","tomato3","red4")))

```

## 5.3 Ratio of Protection Upstream

### 5.3.1 Calculation and Results

Ok so now we tell terra to remove upstream cells that are SPAs in the calculation of upstream cells. The way I implement this is to use the weight argument in the flowAccummulation function (I have tested the behaviour first on a hypothetical 9x9 grid).
```{r, message = FALSE, results = 'hide', fig.height = 9}
## ratio artificial spa
tmp<-spam[[1]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_artif<- 1 - flow_acc_tmp/flow_acc_total 
# clean up
rm(tmp,flow_acc_tmp)

## ratio agricultural spa
tmp<-spam[[2]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_agric<- 1 - flow_acc_tmp/flow_acc_total 
# clean up
rm(tmp,flow_acc_tmp)

## ratio nature spa
tmp<-spam[[3]]
# invert 0 and 1
tmp<-tmp+1
tmp[tmp==2]<-0
# then compute restricted flow accummulation
flow_acc_tmp <- terra::flowAccumulation(flowdir_total,tmp)
# then calculate the ratio
ratio_natur<- 1 - flow_acc_tmp/flow_acc_total
# clean up
rm(tmp,flow_acc_tmp)

## plot results
plot(ratio_artif,colNA="grey40", main = "Ratio of upstream cells covered by Artificial SPA")
plot(ratio_agric,colNA="grey40",main = "Ratio of upstream cells covered by Agricultural SPA")
plot(ratio_natur,colNA="grey40",main = "Ratio of upstream cells covered by Natural SPA")

## save results as tif
rfp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "Projects","Naturregnskap","Flomkontroll","Norway_R","output_maps")
writeRaster(ratio_agric,file.path(rfp,"ratio_agric.tif"),overwrite=T)
writeRaster(ratio_artif,file.path(rfp,"ratio_artif.tif"),overwrite=T)
writeRaster(ratio_natur,file.path(rfp,"ratio_natur.tif"),overwrite=T)

gc()
```

### 5.3.2 Plots for KS Project

Here the ratio plots for our KS cases. Only ratio of upstream cells covered by natural SPA.

```{r, message = FALSE, results = 'hide', fig.height = 7}
# Aurskog Høland
plot(ratio_natur %>% crop(ahs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Aurskog Holand") 
stv<-osm_get_streets_for_KS(ahs,lcm)
plot(stv,add=T)
# Evje and Hornnes
plot(ratio_natur %>% crop(ehs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Evje and Hornnes") 
stv<-osm_get_streets_for_KS(ehs,lcm)
plot(stv,add=T)
# Holtålen
plot(ratio_natur %>% crop(hos,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Holtaalen") 
stv<-osm_get_streets_for_KS(hos,lcm)
plot(stv,add=T)
#Moss
plot(ratio_natur %>% crop(mos,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Moss") 
stv<-osm_get_streets_for_KS(mos,lcm)
plot(stv,add=T)
# Nordreisa
plot(ratio_natur %>% crop(nrs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Nordreisa") 
stv<-osm_get_streets_for_KS(nrs,lcm)
plot(stv,add=T)
# Stad
plot(ratio_natur %>% crop(sds,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Stad") 
stv<-osm_get_streets_for_KS(sds,lcm)
plot(stv,add=T)
# Vardø
plot(ratio_natur %>% crop(vrs,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Vardo") 
stv<-osm_get_streets_for_KS(vrs,lcm)
plot(stv,add=T)
# Vågan
plot(ratio_natur %>% crop(vas,mask=T),colNA="grey40",
     main="Ratio of upstream cells covered by natural SPA, Vaagan") 
stv<-osm_get_streets_for_KS(vas,lcm)
plot(stv,add=T)
```

## 5.4 Service Flow

### 5.4.1 Calculation and Results

Now we have the ratios, so next we can compute for each service demanding area what the respective service flow is. We can make maps of it and also tables.

```{r, message = FALSE,results='hide'}

## serice flow to meet demand by artificial area
flow_d_artif_s_artif<-sdam_artif*ratio_artif
flow_d_artif_s_agric<-sdam_artif*ratio_agric
flow_d_artif_s_natur<-sdam_artif*ratio_natur

## serice flow to meet demand by agricultural area
flow_d_agric_s_artif<-sdam_agric*ratio_artif
flow_d_agric_s_agric<-sdam_agric*ratio_agric
flow_d_agric_s_natur<-sdam_agric*ratio_natur

## serice flow to meet demand by population
flow_d_pop_s_artif<-sdam_pop*ratio_artif
flow_d_pop_s_agric<-sdam_pop*ratio_agric
flow_d_pop_s_natur<-sdam_pop*ratio_natur

## here add the aggregate ones econ and pop, and save those two as rasters.
flow_d_econ <- flow_d_artif_s_artif + flow_d_artif_s_agric + flow_d_artif_s_natur +
  flow_d_agric_s_artif + flow_d_agric_s_agric + flow_d_agric_s_natur

flow_d_pop <- flow_d_pop_s_artif + flow_d_pop_s_agric + flow_d_pop_s_natur

## save results as tif
rfp<-file.path("C:","Users","muk",
               paste0("OneDrive - Statistisk sentralbyr","\U00E5"),
               "Projects","Naturregnskap","Flomkontroll","Norway_R","output_maps")
writeRaster(flow_d_econ,file.path(rfp,"flow_d_econ.tif"),overwrite=T)
writeRaster(flow_d_pop,file.path(rfp,"flow_d_pop.tif"),overwrite=T)

# Make a table of values
flow_df<-as.data.frame(matrix(ncol=4,nrow=3))
colnames(flow_df)<-c("provision_cat","demand_artif","demand_agric","demand_pop")
flow_df$provision_cat<-c("artif","agric","natur")

flow_df$demand_artif[flow_df$provision_cat=="artif"]<-sum(values(flow_d_artif_s_artif),na.rm=T)
flow_df$demand_artif[flow_df$provision_cat=="agric"]<-sum(values(flow_d_artif_s_agric),na.rm=T)
flow_df$demand_artif[flow_df$provision_cat=="natur"]<-sum(values(flow_d_artif_s_natur),na.rm=T)

flow_df$demand_agric[flow_df$provision_cat=="artif"]<-sum(values(flow_d_agric_s_artif),na.rm=T)
flow_df$demand_agric[flow_df$provision_cat=="agric"]<-sum(values(flow_d_agric_s_agric),na.rm=T)
flow_df$demand_agric[flow_df$provision_cat=="natur"]<-sum(values(flow_d_agric_s_natur),na.rm=T)

flow_df$demand_pop[flow_df$provision_cat=="artif"]<-sum(values(flow_d_pop_s_artif),na.rm=T)
flow_df$demand_pop[flow_df$provision_cat=="agric"]<-sum(values(flow_d_pop_s_agric),na.rm=T)
flow_df$demand_pop[flow_df$provision_cat=="natur"]<-sum(values(flow_d_pop_s_natur),na.rm=T)
```
Print the table.
```{r}
kable(flow_df)
```

### 5.4.2 Plots for KS Project

Let's take a look at the flows for our municipalities of interest in the KS project. Only total flow (from all SPAs) to economic assets and to population.

```{r, message = FALSE, results = 'hide', fig.height = 7}
# Aurskog Høland
plot(flow_d_econ %>% crop(ahs,mask=T),colNA="grey40",
     main="Flow to economic assets, Aurskog Holand") 
stv<-osm_get_streets_for_KS(ahs,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(ahs,mask=T),colNA="grey40",
     main="Flow to population, Aurskog Holand") 
stv<-osm_get_streets_for_KS(ahs,lcm)
plot(stv,add=T)
# Evje and Hornnes
plot(flow_d_econ %>% crop(ehs,mask=T),colNA="grey40",
     main="Flow to economic assets, Evje and Hornnes") 
stv<-osm_get_streets_for_KS(ehs,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(ehs,mask=T),colNA="grey40",
     main="Flow to populations, Evje and Hornnes") 
stv<-osm_get_streets_for_KS(ehs,lcm)
plot(stv,add=T)
# Holtålen
plot(flow_d_econ %>% crop(hos,mask=T),colNA="grey40",
     main="Flow to economic assets, Holtaalen") 
stv<-osm_get_streets_for_KS(hos,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(hos,mask=T),colNA="grey40",
     main="Flow to population, Holtaalen") 
stv<-osm_get_streets_for_KS(hos,lcm)
plot(stv,add=T)
#Moss
plot(flow_d_econ %>% crop(mos,mask=T),colNA="grey40",
     main="Flow to economic assets, Moss") 
stv<-osm_get_streets_for_KS(mos,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(mos,mask=T),colNA="grey40",
     main="Flow to population, Moss") 
stv<-osm_get_streets_for_KS(mos,lcm)
plot(stv,add=T)
# Nordreisa
plot(flow_d_econ %>% crop(nrs,mask=T),colNA="grey40",
     main="Flow to economic assetss, Nordreisa") 
stv<-osm_get_streets_for_KS(nrs,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(nrs,mask=T),colNA="grey40",
     main="Flow to population, Nordreisa") 
stv<-osm_get_streets_for_KS(nrs,lcm)
plot(stv,add=T)
# Stad
plot(flow_d_econ %>% crop(sds,mask=T),colNA="grey40",
     main="Flow to economic assetss, Stad") 
stv<-osm_get_streets_for_KS(sds,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(sds,mask=T),colNA="grey40",
     main="Flow to populations, Stad") 
stv<-osm_get_streets_for_KS(sds,lcm)
plot(stv,add=T)
# Vardø
plot(flow_d_econ %>% crop(vrs,mask=T),colNA="grey40",
     main="Flow to economic assets, Vardo") 
stv<-osm_get_streets_for_KS(vrs,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(vrs,mask=T),colNA="grey40",
     main="Flow to population, Vardo") 
stv<-osm_get_streets_for_KS(vrs,lcm)
plot(stv,add=T)
# Vågan
plot(flow_d_econ %>% crop(vas,mask=T),colNA="grey40",
     main="Flow to economic assets, Vaagan") 
stv<-osm_get_streets_for_KS(vas,lcm)
plot(stv,add=T)
plot(flow_d_pop %>% crop(vas,mask=T),colNA="grey40",
     main="Flow to population, Vaagan") 
stv<-osm_get_streets_for_KS(vas,lcm)
plot(stv,add=T)
```

# 6. Met and Unmet Demand

Unmet demand is equal to total demand minus provided service. We can express both as absolute and relative numbers. I make a table for this.

```{r, message = FALSE,results='hide'}

# I make a table of all
dem_df<-as.data.frame(matrix(ncol = 4,nrow=7))
colnames(dem_df)<-c("indicator","demand_artif","demand_agric","demand_pop")
dem_df$indicator<-c("total_demand","total_provided","unmet_demand","rel_prov_artif","rel_prov_agric","rel_prov_natur","rel_unmet")

dem_df$demand_artif[dem_df$indicator=="total_demand"]<-sum(values(sdam_artif),na.rm=T)
dem_df$demand_agric[dem_df$indicator=="total_demand"]<-sum(values(sdam_agric),na.rm=T)
dem_df$demand_pop[dem_df$indicator=="total_demand"]<-sum(values(sdam_pop),na.rm=T)

dem_df$demand_artif[dem_df$indicator=="total_provided"]<-sum(flow_df$demand_artif)
dem_df$demand_agric[dem_df$indicator=="total_provided"]<-sum(flow_df$demand_agric)
dem_df$demand_pop[dem_df$indicator=="total_provided"]<-sum(flow_df$demand_pop)

dem_df$demand_artif[dem_df$indicator=="unmet_demand"]<-dem_df$demand_artif[dem_df$indicator=="total_demand"] - dem_df$demand_artif[dem_df$indicator=="total_provided"]
dem_df$demand_agric[dem_df$indicator=="unmet_demand"]<-dem_df$demand_agric[dem_df$indicator=="total_demand"] - dem_df$demand_agric[dem_df$indicator=="total_provided"]
dem_df$demand_pop[dem_df$indicator=="unmet_demand"]<-dem_df$demand_pop[dem_df$indicator=="total_demand"] - dem_df$demand_pop[dem_df$indicator=="total_provided"]

dem_df$demand_artif[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_artif[flow_df$provision_cat=="artif"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_agric[flow_df$provision_cat=="artif"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_artif"]<-round(flow_df$demand_pop[flow_df$provision_cat=="artif"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)

dem_df$demand_artif[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_artif[flow_df$provision_cat=="agric"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_agric[flow_df$provision_cat=="agric"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_agric"]<-round(flow_df$demand_pop[flow_df$provision_cat=="agric"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)

dem_df$demand_artif[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_artif[flow_df$provision_cat=="natur"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_agric[flow_df$provision_cat=="natur"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_prov_natur"]<-round(flow_df$demand_pop[flow_df$provision_cat=="natur"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)

dem_df$demand_artif[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_artif[dem_df$indicator=="unmet_demand"]/dem_df$demand_artif[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_agric[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_agric[dem_df$indicator=="unmet_demand"]/dem_df$demand_agric[dem_df$indicator=="total_demand"],digits=3)
dem_df$demand_pop[dem_df$indicator=="rel_unmet"]<-round(dem_df$demand_pop[dem_df$indicator=="unmet_demand"]/dem_df$demand_pop[dem_df$indicator=="total_demand"],digits=3)
```
Relevant to note here is that we use population number in case of demand category "pop" and we use cells in case of demand categories "artif" and "agric". We may want to switch all to the count of cells. Then the respective result just has to be multiplied by the spatial size of each cell (each cell one hectare I believe).
```{r}
kable(dem_df)
```

# 7. Supply and Use Tables (SUT)

## 7.1 SUT Norway and NUTS2

First we make a land cover map showing the supply categories we want to report in. And we need to aggregate our flows.
```{r, message = FALSE, results = 'hide', fig.height = 9}

# aggregate flows
flow_d_all_s_all<-flow_d_artif_s_artif+flow_d_artif_s_agric+flow_d_artif_s_natur+
  flow_d_agric_s_artif+flow_d_agric_s_agric+flow_d_agric_s_natur

# so first make the vector of ecosystem types
estv<-c("Settlements and other artificial areas",
        "Cropland",
        "Grassland",
        "Forest and woodlands",
        "Heathlands and shrub",
        "Sparesly vegetated ecosystems",
        "Inland wetlands",
        "Rivers and canals",
        "Lakes and reservoirs",
        "Marine inlets and transitional waters",
        "Coastal beaches, dunes and wetlands",
        "Marine ecosystems")

# and add it to our lcm_idf
lcm_idf['es_type']<-NA
lcm_idf$es_type[lcm_idf$landcover_id %in% c(100,111,112,121,130,131,132,133,134,135,141,142,152)]<-estv[1] # settlem
lcm_idf$es_type[lcm_idf$landcover_id %in% c(200)]<-estv[2] # cropl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(310,320)]<-estv[3] # grassl
lcm_idf$es_type[lcm_idf$landcover_id %in% c(400,410,420,440)]<-estv[4] # forest
lcm_idf$es_type[lcm_idf$landcover_id %in% c(520)]<-estv[5] # bush
lcm_idf$es_type[lcm_idf$landcover_id %in% c(600,611,623,631)]<-estv[6] # sparse
lcm_idf$es_type[lcm_idf$landcover_id %in% c(720)]<-estv[7] # wetlands
lcm_idf$es_type[lcm_idf$landcover_id %in% c(810)]<-estv[8] # rivers
lcm_idf$es_type[lcm_idf$landcover_id %in% c(910)]<-estv[9] # lakes
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1100)]<-estv[11] # coastal
lcm_idf$es_type[lcm_idf$landcover_id %in% c(1200)]<-estv[10] # marine inlets
lcm_idf$es_type[lcm_idf$landcover_id %in% c(9900)]<-estv[12] # marine ecosystems

# Call it lcm5, which follows the required classification
lcm5 <- terra::subst(lcm2,lcm_idf$landcover_id,lcm_idf$es_type)
terra::plot(lcm5, main="Grunnkartet translated to SUT categories")

```

We want to disaggregate the flows by NUTS2 region. So for this we need the individual NUTS2 shapes.
```{r, message = FALSE, results = 'hide', fig.height = 9}
n2shp<-terra::vect(file.path(dfp,"Norway_files","NUTS2021_NO_LVL2_wosvalbard","NUTS2021_NO_LVL2_wos.shp"))
n2shp <- project(n2shp,crs(lcm))
```
Then we compute the flow of service provision for each of the ecosystem categories in each region of interest. We do this by allocating the total service flow across our ecosystems of interest. Allocation follows weights, and the computation of those is specified in the guidance note. I make a function for this, which we can feed with any shapefile of interest for area-specific results.
```{r, message = FALSE,results='hide'}

# here is a function (written for easy debugging, not efficiency)

calc_es_flow_for_shp<-function(estv, # array of unique ecosystems
                               maskshp, # the area of interest as a spatial vector file
                               reflcm, # the reference land cover map
                               totalflowm, # the map of total service flows
                               refcnm # the map of reference curve number
){
  # Step 1: Crop and mask input data (if that is wanted)
  if(is.null(maskshp)){
    cnmc<-refcnm
    lcmc<-reflcm
    flowtc<-totalflowm
  } else {
    cnmc<-crop(refcnm,maskshp,mask=T)
    lcmc<-crop(reflcm,maskshp,mask=T)
    flowtc<-crop(totalflowm,maskshp,mask=T)
  }
  # Step 2: Make a vector of correction factors based on area-specific CN
  cfvc<-numeric()
  for(i in 1:length(estv)){
    cfvc[i]<-(100 - mean(cnmc[lcmc==estv[i]],na.rm=T))/100
  }
  # Step 3: Compute extent of each ecosystem
  exmc<-freq(lcmc)
  # Step 4: Move into one dataframe
  # then put all into one dataframe
  cfmc<-data.frame(cbind(estv,cfvc))
  dfwc<-cfmc %>% left_join(exmc,by=join_by(estv==value)) %>% 
    slice(order(factor(estv, levels = estv))) %>%
    mutate(cfvc = as.numeric(cfvc))
  # Step 5: Compute weighted extents
  dfwc['weighted_ext']<-dfwc$count*dfwc$cfvc
  sswc<-sum(dfwc$weighted_ext,na.rm=T) # sum weighted
  # Step 6: Compute allocation coefficients
  dfwc['alloc_coeff']<-dfwc$weighted_ext/sswc
  # Step 7: Compute total flow for the area of interest
  totalflowc<-sum(values(flowtc),na.rm=T)
  # Step 8: Compute disaggregated flows
  dfwc['fc_flow']<-dfwc$alloc_coeff*totalflowc
  # Step 9: Put all into one table
  dfwtc <- dfwc %>% select(c(estv,fc_flow)) %>% rename(ecosystem = estv) %>%
    add_row(ecosystem = "Total",fc_flow = totalflowc, .before = 1)
  # Step 10: Return result
  return(dfwtc)
}
```
So then we compute our flood protection service flows disaggregated by ecosystem and area. The unit is ha of economic assets protected from potential flood damage. Population is not considered here.
```{r, message = FALSE,results='hide'}
# Loop across our NUTS2 areas
sl<-list()
for(n in 1:length(unique(n2shp$NUTS_ID))){
  sl[[n]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = n2shp[n2shp$NUTS_ID == unique(n2shp$NUTS_ID)[n],],
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
  sl[[n]]['region'] <- unique(n2shp$NUTS_ID)[n]
}
# add for total country
ti<-length(sl)+1
sl[[ti]]<-calc_es_flow_for_shp(estv=estv,
                                         maskshp = NULL,
                                         reflcm = lcm5,
                                         totalflowm = flow_d_all_s_all,
                                         refcnm = cnm2[[4]])
sl[[ti]]['region'] <- "NO0"
# reformat to our final supply table
sdf<-do.call("rbind",sl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)
library(tidyr)
rm(sl,ti)
```
Here is the resulting supply table:
```{r}
kable(sdf)
```
And then the use table. For the time being we only report total use and we assume it goes to households.
```{r, message = FALSE,results='hide'}
udf <- sdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
```
Here is the use table:
```{r}
kable(udf)
```

## 7.2 SUT KS Project

And here for our case study municipalities
```{r}
# Loop across our NUTS2 areas
kl<-list()
kl[[1]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ahs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[1]]['region'] <- "Aurskog-Holand" 
kl[[2]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = ehs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[2]]['region'] <- "Evje og Hornnes" 
kl[[3]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = hos,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[3]]['region'] <- "Holtaalen" 
kl[[4]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = mos,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[4]]['region'] <- "Moss" 
kl[[5]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = nrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[5]]['region'] <- "Nordreisa" 
kl[[6]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = sds,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[6]]['region'] <- "Stad" 
kl[[7]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vrs,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[7]]['region'] <- "Vardo" 
kl[[8]]<-calc_es_flow_for_shp(estv=estv,
                                maskshp = vas,
                                reflcm = lcm5,
                                totalflowm = flow_d_all_s_all,
                                refcnm = cnm2[[4]])
kl[[8]]['region'] <- "Vaagan" 
# reformat to our final supply table
kdf<-do.call("rbind",kl) %>% pivot_wider(names_from = ecosystem,values_from = fc_flow) %>% arrange(region)

# create use table
kudf <- kdf %>% 
  select(c(region,Total)) %>% 
  mutate(`Intermediate consumption by industries` = NA) %>%
  mutate(`Government final consumption` = NA) %>%
  mutate(`Households final consumption` = Total) %>%
  mutate(`Gross capital formation` = NA) %>%
  mutate(Exports = NA)
```
Here is the supply table:
```{r}
kable(kdf)
```
Here is the use table:
```{r}
kable(kudf)
```


# 8. Ideas for Improvements

-   Add street infrastructure from OSM to our artificial areas
-   In section 3.2.2 we could revisit the hydrological soil type map, possibly using a more specific map for Norway. The one we are currently using has missing data in the North of Norway
-   In section 3.2.3 we could revisit the CN lookup table. In its current form there are some categories missing, which means that our resulting potential runoff retention map has a quite a few "holes", i.e. cells that cannot provide any flood control in our model, even though they probably do in reality. These categories are 9900 (uklassifisert), 1200 (marine økosystemer (uspesifisert)), 1100 (Kyststrender dyner og våtmark (uspesifisert)), 910 (Innsjøer og dammer), 810 (Elver og bekker), and 631 (Isbreer og varige snøfonner).
-   In section 3.5 we could revisit the threshold definition for SPAs. While the current definition is clear and transparent, its scientific substance is unclear.
-   In section 4 we could refine service demand, for example by changing to a different flood frequency or/and a different flood depth threshold.
-   In section 7 we could provide results by catchment. This is something the guidance note suggests and it should be easy to implement.
